---
title: 'Aggregate IUCN spp ranges, range-rarity weighted'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/src/templates/ohara_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(raster)
library(data.table)
source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')  ###
  ### includes library(tidyverse); library(stringr); dir_M points to ohi directory
library(sf)

dir_git <- here()

source(file.path(dir_git, '_setup/common_fxns.R'))

```

# Summary

Using individual species range data (csv outputs from `5_process_spp_shps.Rmd`), collect into species groups.  In each species group, process range and risk information to get mean risk, variance of risk, species richness, and threatened species count per cell.  These are all weighted by species range rarity as a metric of endemism.

# Methods

## Generate species group maps - range-rarity weighting per cell

* Determine species ranges using ocean-area data.frame and cell location data.frame.  Create an overall data.frame and save as .csv.
    * Compare to polygon range areas
* Collect all taxon species into a single data.frame.
* Join species extinction risk and trend data.frame to species-cell data.frame, including regional assessments.
* Join species range area 
* By cell, calculate range-rarity-weighted mean extinction risk, variance of extinction risk, trend, threatened species count, rr_richness, rr_trend_richness (where rr_richness is range-rarity weighted species richness)
    * When aggregating further, use rr_richness to weight values.
    
__NOTE:__ as above, because the summary files are likely to be very large for globe-spanning taxa, save these outputs outside of GitHub.

### Calculate range from the rasterized range maps

This calculated range includes ONLY ocean area. For species such as birds whose range is primarily over land but extends somewhat out over the ocean, this will underpredict range according to the polygons.  For marine species whose range is extremely tiny, the raster may miss the presence of a species entirely.  These ranges will be checked against the polygon areas as a quality check.

``` {r calculate_ranges_from_raster_maps}

spp_range_file <- file.path(dir_data,
                            sprintf('iucn_spp_range_area_%s.csv', api_version))
reload <- FALSE

### load area raster, convert to data frame
area_rast <- raster(file.path(dir_spatial, 'ocean_area_rast.tif'))
area_df <- data.frame(cell_id = 1:length(area_rast),
                      area_km2 = values(area_rast))
  
### load csvs of marine species range maps
spp_maps <- read_csv(file.path(dir_data, 
                               sprintf('spp_marine_maps_%s.csv', api_version)),
                     col_types = 'ddcicccl')
spp_ids <- spp_maps$iucn_sid %>%
  unique() %>%
  sort()

if(reload == FALSE & file.exists(spp_range_file)) {
  ### don't redo spp already calculated...  remove those from the running
  spp_range_list_done <- read_csv(spp_range_file, col_types = 'ddcc')
  spp_ids <- spp_ids[!spp_ids %in% spp_range_list_done$iucn_sid]
}

message('Processing ', length(spp_ids), ' species range maps.')
  
if(length(spp_ids) > 0) { ### any more spp to calculate?
  system.time({
  spp_range_list <- parallel::mclapply(seq_along(spp_ids), mc.cores = 24,
                                       FUN = function(i) {
    x <- spp_ids[i]
    cat_msg(i, ': Processing species ', x)
    spp_map <- read_csv(file.path(dir_o_anx, 'spp_rasters',
                                  sprintf('iucn_sid_%s.csv', x)),
                        col_types = 'di')
    
    spp_map_area <- spp_map %>% 
      # filter(presence != 5) %>% ### exclude extinct areas
      inner_join(area_df, by = 'cell_id')
    
    spp_range <- data.frame(iucn_sid  = x,
                            range_km2 = sum(spp_map_area$area_km2, na.rm = TRUE))
    # spp_range_list[[i]] <- spp_range
    return(spp_range)
  # }
  }) ### end of mclapply
  }) ### end of system.time

  spp_range_df <- bind_rows(spp_range_list) %>%
    left_join(spp_maps %>% 
                select(iucn_sid, sciname, dbf_file) %>%
                distinct(),
              by = 'iucn_sid')
} else {
  spp_range_df <- data.frame() ### return an empty data.frame
}
  
if(reload == FALSE & exists('spp_range_list_done')) {
  ### add new area calcs at the end of the dataframe of spp already done
  spp_range_df <- spp_range_df %>%
    bind_rows(spp_range_list_done)
}

write_csv(spp_range_df, spp_range_file)

```

### calculate range from polygons

This may *over*predict range for many species whose range is limited by depth,
due to polyon buffering.  This can be used to sanity-check the 
raster-calculated ranges.

``` {r calculate range from polygons, eval = TRUE}

spp_maps <- read_csv(file.path(dir_data, sprintf('spp_marine_maps_%s.csv', api_version)),
                     col_types = 'ddcicccl')

spp_dbfs <- spp_maps$dbf_file %>%
  unique() 
spp_shps <- spp_dbfs %>%
  str_replace('\\.dbf$', '.shp')

polygon_areas_file <- file.path(dir_setup, 'int/polygon_areas.csv')

reload <- FALSE
if(!file.exists(polygon_areas_file) | reload) {
  
  poly_range_list <- parallel::mclapply(seq_along(spp_shps), mc.cores = 36,
                          FUN = function(i) { ### i <- 5
                            x <- spp_shps[i]
                            cat_msg(i, ': Processing ', x)
                            shp <- st_read(x)
                            
                            if('id_no' %in% names(shp)) {
                              shp <- shp %>%
                                rename(iucn_sid = id_no)
                            }
                            
                            shp <- shp %>%
                              filter(iucn_sid %in% spp_maps$iucn_sid)
                                ### NOTE: this will miss subpops

                            range_df <- data.frame(iucn_sid = shp$iucn_sid,
                                                   poly_range_km2 = st_area(shp)) %>%
                              mutate(poly_range_km2 = as.numeric(poly_range_km2 / 1e6))
                            cat_msg(i, ': Finished ', x)
                            
                            return(range_df)
                          })
  
  poly_range_df <- poly_range_list %>%
    setNames(basename(spp_shps)) %>%
    bind_rows(.id = 'shp_file') %>%
    group_by(iucn_sid) %>%
    summarize(poly_range_km2 = round(max(poly_range_km2), 3))

  write_csv(poly_range_df, polygon_areas_file)
}
    
```

### plot raster-calculated ranges vs polygon-calculated ranges

``` {r, eval = TRUE}

range_df <- read_csv(polygon_areas_file, col_types = 'dd') %>%
  full_join(read_csv(spp_range_file, col_types = 'ddcc') %>%
              mutate(shp_file = str_replace(dbf_file, 'dbf', 'shp')), 
            by = 'iucn_sid') %>%
  mutate(birds = str_detect(shp_file, 'bli'))

range_compare_scatter <- ggplot(range_df, aes(x = range_km2, y = poly_range_km2)) +
  ggtheme_plot() +
  geom_abline(slope = 1, intercept = 0, color = 'green4') + 
  geom_point(aes(color = birds, label = shp_file, label2 = iucn_sid), alpha = .6)

print(range_compare_scatter)
```

Where dots fall on the line, species ranges by raster and by polygon agree.  Where dots fall above the line, polygon range is larger than raster range - likely due to clipping of land-based ranges, and mostly birds.  Where dots fall below the line, polygon range is smaller than the raster range - in these cases, the polygon range may be a combination of multiple polygons, e.g. subpops or seasonal areas; the code above simply takes the largest polygon per ID as a rough check.

## Aggregate species maps to taxa level

Species cell values for category and trend are multiplied by range rarity before summarizing to mean and variance.  Rather than species counts, the sum of range rarity is calculated for each cell.

Since most of these values will be tiny (since range rarity is generally a tiny number for species with any significant area), we can later multiply these by a scaling factor to get something more in line with more convenient orders of magnitude.

``` {r aggregate species maps to taxa}

### Read in lots of data
spp_maps <- read_csv(file.path(dir_data, sprintf('spp_marine_maps_%s.csv', api_version)),
                     col_types = 'ddcicccl')

spp_risk <- read_csv(file.path(dir_data, sprintf('iucn_risk_current_%s.csv', api_version)),
                     col_types = 'dccicccdc')
spp_risk_rgn <- read_csv(file.path(dir_data, sprintf('iucn_risk_rgn_current_%s.csv', api_version)),
                         col_types = 'dcccdc')

spp_trend <- read_csv(file.path(dir_data, sprintf('iucn_trend_by_spp_%s.csv', api_version)),
                      col_types = 'dc_d_')

### load range info and ocean area info
spp_range <- read_csv(file.path(dir_data,
                                sprintf('iucn_spp_range_area_%s.csv', api_version)),
                      col_types = 'dd__')

# ocean_area_rast <- raster(file.path(dir_spatial, 'ocean_area_rast.tif'))
# cell_area_df <- data.frame(cell_ocean_area = values(ocean_area_rast),
#                            cell_id = 1:length(ocean_area_rast))

### make a dataframe of species risk, trend, regional risk, and species range
spp_risk_trend <- spp_risk %>%
  mutate(iucn_rgn = 'global') %>%
  bind_rows(spp_risk_rgn) %>%
  select(iucn_sid, iucn_rgn, cat_score) %>%
  left_join(spp_trend, by = c('iucn_sid', 'iucn_rgn')) %>%
  left_join(spp_range, by = 'iucn_sid')

### raster for cell IDs
rast_cell_ids <- raster(file.path(dir_spatial, 'cell_id_rast.tif'))

### Make a dataframe of cell ID to MEOW for regional assessments... also 
### make a lookup of MEOW to region
meow_rgns_rast <- raster::raster(file.path(dir_spatial, 'meow_rast.tif'))
meow_cells_all <- data.frame(cell_id = values(rast_cell_ids),
                             meow_id = values(meow_rgns_rast)) %>%
  filter(!is.na(meow_id))
meow_to_rgn <- read_csv(file.path(dir_spatial, 'iucn_rgn_to_meow.csv'))


```

## Loop over taxonomic groups

``` {r loop over comp-assessed taxa}

### Make a list of taxonomic groups to loop over:
taxa <- spp_maps %>%
  filter(comp_assessed == TRUE) %>%
  .$dbf_file %>%
  basename() %>% unique() %>%
  str_replace('\\.dbf$', '')

# taxa <- taxa[str_detect(taxa, 'acipenser|bli_marine|TERR|SEAGRASS')]

reload <- FALSE

for(taxon in taxa) {
  ### taxon <- 'SEAGRASSES'
  ### taxon <- taxa[1]
  
  taxon_risk_sum_file <- file.path(dir_o_anx, 'taxa_summaries',
                                   sprintf('%s_cell_sum_comp_rrweight_%s.csv', 
                                           tolower(taxon), api_version))
  
  if(!file.exists(taxon_risk_sum_file) | reload == TRUE) {
    
    taxon_maps <- spp_maps %>%
      filter(str_detect(dbf_file, taxon))
    
    taxon_risk_trend <- spp_risk_trend %>%
      filter(iucn_sid %in% taxon_maps$iucn_sid) %>%
      filter(!is.na(cat_score)) %>%
      arrange(iucn_sid)
    
    ### Using the iucn_sid field, generate a vector of all species range files for
    ### this taxon.
    taxon_ids <- taxon_risk_trend$iucn_sid %>%
      unique()
    
    cat_msg('Processing ', length(taxon_ids), ' species maps in ', basename(taxon), '...')

    ##########################################################.
    ### Looping over species within group -----
    ##########################################################.
    ### Collect all species ranges for this taxon into a single data.frame.
    ### Use mclapply since we're reading many large-ish files.  For MARINE_MAMMALS (85 assessed spp)
    ### this takes about 30-40 seconds
    taxon_cells_list <- parallel::mclapply(taxon_ids, mc.cores = 24,
                                        FUN = function(x) {
        ### x <- taxon_ids[11]
        ### x <- 6336                            
        csv_file <- file.path(dir_o_anx, 'spp_rasters', 
                              sprintf('iucn_sid_%s.csv', x))
        
        spp_risk_map <- read_csv(csv_file, col_types = 'di') %>%
          mutate(iucn_sid = x) %>%
          left_join(taxon_risk_trend, by = 'iucn_sid') 
        
        ### Identify regional assessments if any
        meow_rgns <- meow_to_rgn %>%
          filter(iucn_rgn %in% spp_risk_map$iucn_rgn)
        
        non_global_rgns <- meow_rgns %>%
          filter(iucn_rgn != 'global')
        
        if(nrow(non_global_rgns) > 0) {
          ### If any regional assessments, clip the MEOW cells down to the appropriate region...
          meow_cells <- meow_cells_all %>%
            inner_join(meow_rgns, by = 'meow_id') %>%
            rename(rgn_name = iucn_rgn)
          ### ... then filter out non-matching overlapped cells
          spp_risk_map <- spp_risk_map %>%
            left_join(meow_cells, by = 'cell_id') %>%
            mutate(rgn_name = ifelse(is.na(meow_id), 'global', rgn_name),
                   priority = ifelse(rgn_name == 'global', 100, priority)) %>%
            filter(iucn_rgn == rgn_name) %>%
            group_by(cell_id) %>%
            filter(priority == min(priority)) %>%
            ungroup()
          ### NOTE: at this point, still possible to have multiple regional
          ### assessments, if priorities are the same (e.g. Europe and Pan Africa,
          ### or overlaps around Africa).  Those values will be averaged in
          ### the group_by() below.
        }
        
        ### calc range rarity and select down to main columns; also, 
        ### if presence == 5 (extinct), adjust category and trend scores.
        spp_risk_map <- spp_risk_map %>%
          # left_join(cell_area_df, by = 'cell_id') %>%
          select(cell_id, presence, iucn_sid, iucn_rgn, 
                 cat_score, trend_score, range_km2) %>% # , cell_ocean_area) %>%
          distinct() %>%
          mutate(range_rarity = ifelse(range_km2 > 0, 1 / range_km2, 0),
                 cat_score   = ifelse(presence == 5, 1, cat_score),
                 trend_score = ifelse(presence == 5, NA, trend_score)) %>%
          select(-range_km2) %>% # , cell_ocean_area) %>%
            ### drop variables no longer needed
          group_by(cell_id, iucn_sid) %>%
          summarize(cat_score   = mean(cat_score, na.rm = TRUE),
                    trend_score = mean(trend_score, na.rm = TRUE),
                    iucn_rgn = paste0(iucn_rgn, collapse = ','),
                    presence = paste0(presence, collapse = ','),
                    range_rarity = first(range_rarity)) %>%
          ungroup() %>%
          mutate(trend_score = ifelse(is.nan(trend_score), NA, trend_score))
        
        return(spp_risk_map)
      }) ### end of mclapply over all species in taxonomic group
  
    ##########################################################.
    ### Processing cell calculations for group -----
    ##########################################################.
  
    ### Set up for keyed data.table merging: key for iucn_sid, cell_id
    cat_msg('...binding rows into data.frame...')
    taxon_risk_map <- bind_rows(taxon_cells_list)
    
    cat_msg('...summarizing...')
    taxon_risk_summary <- taxon_risk_map %>%
      group_by(cell_id) %>%
      summarize(mean_risk = sum(cat_score * range_rarity) / sum(range_rarity), ### NA categories already filtered out
                sr_rr_risk  = sum(range_rarity),
                v1 = sum(range_rarity[!is.na(cat_score)]),   ### for recombining group variances
                v2 = sum(range_rarity[!is.na(cat_score)]^2), ### for recombining group variances
                alpha     = v1 - (v2 / v1), 
                var_risk  = sum(range_rarity * (cat_score - mean_risk)^2) / alpha,
                  ### see https://en.wikipedia.org/wiki/Weighted_arithmetic_mean#Weighted_sample_variance
                var_risk  = ifelse(is.nan(var_risk), NA, var_risk),
                  ### for cells with only one spp in the cell, returns NaN, reset to NA
                sr_rr_threatened = sum((cat_score >= 0.4 & cat_score < 1) * range_rarity),  
                mean_trend  = sum(trend_score * range_rarity, na.rm = TRUE) /
                  sum(range_rarity[!is.na(trend_score)]),
                mean_trend  = ifelse(is.nan(mean_trend), NA, mean_trend),
                sr_rr_trend = sum(range_rarity[!is.na(trend_score)]),
                n_spp = n()) %>%
      select(-alpha)
    
    cat_msg('...writing file', taxon_risk_sum_file, '...')
    write_csv(taxon_risk_summary, taxon_risk_sum_file)

  } else { ### end of if statement checking whether file exists for this taxon
    cat_msg('Found file ', taxon_risk_sum_file, '... skipping process...')
  }
  
} ### end of taxonomic group loop
  
```

``` {r loop over non-comp-assessed taxa}

### Make a list of taxonomic groups to loop over:
taxa <- spp_maps %>%
  filter(comp_assessed == FALSE) %>%
  .$dbf_file %>%
  basename() %>% unique() %>%
  str_replace('\\.dbf$', '')
# taxa <- taxa[c(5, 22, 13, 25)]

for(taxon in taxa) {
  ### taxon <- 'SEAGRASSES'
  ### taxon <- taxa[1]
  
  taxon_risk_sum_file <- file.path(dir_o_anx, 'taxa_summaries',
                                   sprintf('%s_cell_sum_noncomp_rrweight_%s.csv', 
                                           tolower(taxon), api_version))
  
  reload <- FALSE
  if(!file.exists(taxon_risk_sum_file) | reload == TRUE) {
    
    taxon_maps <- spp_maps %>%
      filter(str_detect(dbf_file, taxon)) %>%
      filter(comp_assessed == FALSE) %>%
      filter(max_depth != 'deep oceanic')
    
    taxon_risk_trend <- spp_risk_trend %>%
      filter(iucn_sid %in% taxon_maps$iucn_sid) %>%
      filter(!is.na(cat_score)) %>%
      arrange(iucn_sid)
    
    if(nrow(taxon_risk_trend) == 0) {
      ### if empty set, create a dummy data frame so there
      ### will at least be a file...
      cat_msg('Creating dummy data frame for empty spp set ', taxon)
      taxon_risk_summary <- data.frame(
        cell_id = -1,
        mean_risk = NA)
    } else {
      ### Using the iucn_sid field, generate a vector of all species range files for
      ### this taxon.
      taxon_ids <- taxon_risk_trend$iucn_sid %>%
        unique()
      
      cat_msg('Processing ', length(taxon_ids), ' species maps in ', basename(taxon), '...')
  
      ##########################################################.
      ### Looping over species within group -----
      ##########################################################.
      ### Collect all species ranges for this taxon into a single data.frame.
      ### Use mclapply since we're reading many large-ish files.  For MARINE_MAMMALS (85 assessed spp)
      ### this takes about 30-40 seconds
      taxon_cells_list <- parallel::mclapply(taxon_ids, mc.cores = 24,
                                          FUN = function(x) {
          ### x <- taxon_ids[11]
          ### x <- 6336                            
          csv_file <- file.path(dir_o_anx, 'spp_rasters', 
                                sprintf('iucn_sid_%s.csv', x))
          
          spp_risk_map <- read_csv(csv_file, col_types = 'di') %>%
            mutate(iucn_sid = x) %>%
            left_join(taxon_risk_trend, by = 'iucn_sid') 
          
          ### Identify regional assessments if any
          meow_rgns <- meow_to_rgn %>%
            filter(iucn_rgn %in% spp_risk_map$iucn_rgn)
          
          non_global_rgns <- meow_rgns %>%
            filter(iucn_rgn != 'global')
          
          if(nrow(non_global_rgns) > 0) {
            ### If any regional assessments, clip the MEOW cells down to the appropriate region...
            meow_cells <- meow_cells_all %>%
              inner_join(meow_rgns, by = 'meow_id') %>%
              rename(rgn_name = iucn_rgn)
            ### ... then filter out non-matching overlapped cells
            spp_risk_map <- spp_risk_map %>%
              left_join(meow_cells, by = 'cell_id') %>%
              mutate(rgn_name = ifelse(is.na(meow_id), 'global', rgn_name),
                     priority = ifelse(rgn_name == 'global', 100, priority)) %>%
              filter(iucn_rgn == rgn_name) %>%
              group_by(cell_id) %>%
              filter(priority == min(priority)) %>%
              ungroup()
            ### NOTE: at this point, still possible to have multiple regional
            ### assessments, if priorities are the same (e.g. Europe and Pan Africa,
            ### or overlaps around Africa).  Those values will be averaged in
            ### the group_by() below.
          }
          
          ### calc range rarity and select down to main columns; also, 
          ### if presence == 5 (extinct), adjust category and trend scores.
          spp_risk_map <- spp_risk_map %>%
            # left_join(cell_area_df, by = 'cell_id') %>%
            select(cell_id, presence, iucn_sid, iucn_rgn, 
                   cat_score, trend_score, range_km2) %>% # , cell_ocean_area) %>%
            distinct() %>%
            mutate(range_rarity = ifelse(range_km2 > 0, 1 / range_km2, 0),
                   cat_score   = ifelse(presence == 5, 1, cat_score),
                   trend_score = ifelse(presence == 5, NA, trend_score)) %>%
            select(-range_km2) %>% # , cell_ocean_area) %>%
              ### drop variables no longer needed
            group_by(cell_id, iucn_sid) %>%
            summarize(cat_score   = mean(cat_score, na.rm = TRUE),
                      trend_score = mean(trend_score, na.rm = TRUE),
                      iucn_rgn = paste0(iucn_rgn, collapse = ','),
                      presence = paste0(presence, collapse = ','),
                      range_rarity = first(range_rarity)) %>%
            ungroup() %>%
            mutate(trend_score = ifelse(is.nan(trend_score), NA, trend_score))
          
          return(spp_risk_map)
        }) ### end of mclapply over all species in taxonomic group
    
      ##########################################################.
      ### Processing cell calculations for group -----
      ##########################################################.
    
      ### Set up for keyed data.table merging: key for iucn_sid, cell_id
      cat_msg('...binding rows into data.frame...')
      taxon_risk_map <- bind_rows(taxon_cells_list)
      
      cat_msg('...summarizing...')
      taxon_risk_summary <- taxon_risk_map %>%
        group_by(cell_id) %>%
        summarize(mean_risk = sum(cat_score * range_rarity) / sum(range_rarity), ### NA categories already filtered out
                  sr_rr_risk  = sum(range_rarity),
                  v1 = sum(range_rarity[!is.na(cat_score)]),   ### for recombining group variances
                  v2 = sum(range_rarity[!is.na(cat_score)]^2), ### for recombining group variances
                  alpha     = v1 - (v2 / v1), 
                  var_risk  = sum(range_rarity * (cat_score - mean_risk)^2) / alpha,
                    ### see https://en.wikipedia.org/wiki/Weighted_arithmetic_mean#Weighted_sample_variance
                  var_risk  = ifelse(is.nan(var_risk), NA, var_risk),
                    ### for cells with only one spp in the cell, returns NaN, reset to NA
                  sr_rr_threatened = sum((cat_score >= 0.4 & cat_score < 1) * range_rarity),  
                  mean_trend  = sum(trend_score * range_rarity, na.rm = TRUE) /
                    sum(range_rarity[!is.na(trend_score)]),
                  mean_trend  = ifelse(is.nan(mean_trend), NA, mean_trend),
                  sr_rr_trend = sum(range_rarity[!is.na(trend_score)]),
                  n_spp = n()) %>%
        select(-alpha)
    } ### end of empty set check
    
    cat_msg('...writing file', taxon_risk_sum_file, '...')
    write_csv(taxon_risk_summary, taxon_risk_sum_file)

  } else { ### end of if statement checking whether file exists for this taxon
    cat_msg('Found file ', taxon_risk_sum_file, '... skipping process...')
  }
  
} ### end of taxonomic group loop
  
```
