---
title: 'Biodiversity risk covariance exploration'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohibc/src/templates/ohibc_hdr1.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(raster)
library(data.table)

source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')
  ### includes library(tidyverse); library(stringr); 
  ### dir_M points to ohi directory on Mazu; dir_O points to home dir on Mazu

dir_git <- '~/github/spp_risk_dists'

### goal specific folders and info
dir_setup <- file.path(dir_git, 'data_setup')
dir_data  <- file.path(dir_git, 'data')
dir_o_anx <- file.path(dir_O, 'git-annex/spp_risk_dists')

source(file.path(dir_git, 'data_setup/api_fxns.R'))

```

# Summary

Compare MPAs to biodiversity intactness: 

# Methods

## set up taxonomic groups

``` {r set up taxa groupings}

shp_to_taxa <- read_csv(file.path(dir_git, 'data_setup/raw', 'shps_to_taxa.csv'))

taxa_sums <- data.frame(unweighted = list.files(file.path(dir_o_anx, 'taxa_summaries'),
                                                pattern = sprintf('cell_sum_%s.csv', api_version),
                                                full.names = TRUE),
                        rr_weighted = list.files(file.path(dir_o_anx, 'taxa_summaries'),
                                                pattern = sprintf('cell_sum_rrweight_%s.csv', api_version),
                                                full.names = TRUE)) %>%
  mutate(shp_basename = str_replace_all(basename(unweighted), '_cell_sum.+|_part.+', '')) %>%
  left_join(shp_to_taxa, by = 'shp_basename')


# spp_taxonomy <- read_csv(file.path(dir_o_anx, sprintf('iucn/spp_info_from_api_%s.csv', api_version))) %>%
#   select(iucn_sid, sciname, kingdom, phylum, class)
# 
# spp_info <- read_csv(file.path(dir_git, 'data', sprintf('spp_marine_maps_%s.csv', api_version)),
#                      col_types = 'ddciccc') %>% 
#   select(iucn_sid, sciname, dbf_file) %>%
#   left_join(spp_taxonomy %>% select(-sciname), by = 'iucn_sid')

  
# phyla_summaries <- spp_info %>%
#   mutate(phylum = tolower(phylum), class = tolower(class)) %>%
#   group_by(phylum) %>%
#   mutate(n_phylum = n()) %>%
#   group_by(phylum, class, n_phylum) %>%
#   summarize(n_class = n()) %>%
#   ungroup() %>%
#   mutate(taxa_gp = ifelse(phylum == 'chordata', 
#                           paste0('Chordata: ', tools::toTitleCase(class)), 
#                           tools::toTitleCase(phylum)),
#          n_gp = ifelse(phylum == 'chordata', n_class, n_phylum)) %>%
#   select(taxa_gp, n_gp) %>%
#   distinct()


# spp_info_all_marine <- read_csv(file.path(dir_git, 'data', 
#                                           sprintf('spp_marine_from_api_%s.csv', api_version))) %>%
#   left_join(spp_taxonomy, by = 'iucn_sid')
# 
# spp_info_all_marine$class %>% unique()
#  [1] "ACTINOPTERYGII"     "REPTILIA"           "MAMMALIA"           "CHONDRICHTHYES"    
#  [5] "GASTROPODA"         "MAXILLOPODA"        "ENOPLA"             "CEPHALASPIDOMORPHI"
#  [9] "SARCOPTERYGII"      "MEROSTOMATA"        "AMPHIBIA"           "INSECTA"           
# [13] "MAGNOLIOPSIDA"      "PINOPSIDA"          "LILIOPSIDA"         "CYCADOPSIDA"       
# [17] "FLORIDEOPHYCEAE"    "ANTHOZOA"           "ULVOPHYCEAE"        "PHAEOPHYCEAE"      
# [21] "HYDROZOA"           "MALACOSTRACA"       "MYXINI"             "BIVALVIA"          
# [25] "CEPHALOPODA"        "POLYPODIOPSIDA"     "HOLOTHUROIDEA"      "AVES"              
# [29] "LYCOPODIOPSIDA"     "CLITELLATA" 
```

## All spp

### covariance of unweighted vs rr-weighted

Covariance calculated for risk, percent threatened, and species richness.  Does range-rarity weighting provide a significantly different picture of status compared to unweighted?

Here we can use the overall rasters for simplicity.

``` {r covariance checks global, results = 'asis'}

risk_un_rast    <- raster(file.path(dir_git, 'output', 'mean_risk_raster_010deg.tif'))
risk_rr_rast    <- raster(file.path(dir_git, 'output', 'mean_rr_risk_raster_010deg.tif'))
threat_un_rast  <- raster(file.path(dir_git, 'output', 'pct_threat_raster_010deg.tif'))
threat_rr_rast  <- raster(file.path(dir_git, 'output', 'sr_rr_pct_threat_raster_010deg.tif'))
n_spp_rast <- raster(file.path(dir_git, 'output', 'n_spp_risk_raster_010deg.tif'))
n_rr_rast  <- raster(file.path(dir_git, 'output', 'sr_rr_risk_raster_010deg.tif'))

df <- data.frame(risk_unweight = values(risk_un_rast),
                 risk_rrweight = values(risk_rr_rast),
                 threat_unweight = values(threat_un_rast),
                 threat_rrweight = values(threat_rr_rast),
                 nspp_unweight = values(n_spp_rast),
                 nspp_rrweight = values(n_rr_rast)) %>%
  mutate(pct_thr_unweight = threat_unweight / nspp_unweight,
         pct_thr_rrweight = threat_rrweight / nspp_rrweight) %>%
  filter(!is.na(risk_unweight))


mdl_summary <- function(mdl) {
  mdl %>%
    broom::tidy() %>%
    mutate(adj.rsq = round(mdl %>% summary() %>% .$adj.r.squared, 4),
           call = as.character(mdl$call)[2])
}

mdl_summary(lm(risk_rrweight ~ risk_unweight, data = df)) %>%
  knitr::kable()

mdl_summary(lm(threat_rrweight ~ threat_unweight, data = df)) %>%
  knitr::kable()

mdl_summary(lm(nspp_rrweight ~ nspp_unweight, data = df)) %>%
  knitr::kable()

```

### covariance of factors within weighting groups

Compare correlations between:

* species richness ~ n threatened 
* mean risk ~ pct threatened 
* mean risk ~ species richness

``` {r covariance checks global 2, results = 'asis'}

mdl_summary(lm(nspp_unweight ~ threat_unweight, data = df)) %>%
  knitr::kable()

mdl_summary(lm(nspp_rrweight ~ threat_rrweight, data = df)) %>%
  knitr::kable()


mdl_summary(lm(risk_unweight ~ pct_thr_unweight, data = df)) %>%
  knitr::kable()

mdl_summary(lm(risk_rrweight ~ pct_thr_rrweight, data = df)) %>%
  knitr::kable()


mdl_summary(lm(risk_unweight ~ nspp_unweight, data = df)) %>%
  knitr::kable()

mdl_summary(lm(risk_rrweight ~ nspp_rrweight, data = df)) %>%
  knitr::kable()

```

## By taxa: covariance of unweighted vs rr-weighted

Covariance calculated for risk, percent threatened, and species richness.

``` {r calc covar by taxon, results = 'asis'}

spp_info <- read_csv(file.path(dir_git, 'data', sprintf('spp_marine_maps_%s.csv', api_version)),
                     col_types = 'ddciccc') %>%
  mutate(shp_basename = str_replace_all(basename(tolower(dbf_file)), '.dbf|_part.+', '')) %>%
  select(iucn_sid, sciname, shp_basename) %>%
  left_join(taxa_sums, by = 'shp_basename')

taxa_gps <- spp_info$taxon %>% unique()

reload <- FALSE

for(taxon_gp in taxa_gps) { ### taxon_gp <- taxa_gps[1]
  
  taxon_txt <- taxon_gp %>% tolower() %>% str_replace_all('chordata: |[^a-z]+', '')

  taxa_info <- spp_info %>%
    filter(taxon == taxon_gp)
  
  n_spp_in_taxa <- unique(taxa_info$iucn_sid) %>% length()
  	
  message('Processing ', taxon_gp)
  
  unweighted_files <- taxa_info$unweighted %>% unique()
  rr_weighted_files <- taxa_info$rr_weighted %>% unique()
  
  taxon_df <- parallel::mclapply(unweighted_files, mc.cores = 12,
      FUN = function (x) { ### x <- unweighted_files[1]
        read_csv(x, col_types = 'dddiidi')
      }) %>% 
    bind_rows() %>%
    group_by(cell_id) %>%
    summarize(risk_unweight = 1/sum(n_spp_risk) * sum(mean_risk * n_spp_risk),
              threat_unweight = sum(n_spp_threatened) / sum(n_spp_risk),
              nspp_unweight = sum(n_spp_risk))
  
  taxon_df_rr <- parallel::mclapply(rr_weighted_files, mc.cores = 12,
      FUN = function (x) { ### x <- rr_weighted_files[1]
        read_csv(x, col_types = 'ddd___d___')
      }) %>% 
    bind_rows() %>%
    group_by(cell_id) %>%
    summarize(risk_rrweight = 1/sum(sr_rr_risk) * sum(mean_risk * sr_rr_risk),
              threat_rrweight = sum(sr_rr_threatened) / sum(sr_rr_risk),
              nspp_rrweight = sum(sr_rr_risk))

  df <- taxon_df %>%
    left_join(taxon_df_rr, by = 'cell_id')
  
  cat(paste0('\n\n### ', taxon_gp, ' (', n_spp_in_taxa, ' species)\n\n'))

  mdl_summary(lm(risk_rrweight ~ risk_unweight, data = df)) %>%
    knitr::kable() %>%
    print()
  
  mdl_summary(lm(threat_rrweight ~ threat_unweight, data = df)) %>%
    knitr::kable() %>%
    print()
  
  mdl_summary(lm(nspp_rrweight ~ nspp_unweight, data = df)) %>%
    knitr::kable() %>%
    print()

}

```

