---
title: 'MPAs and biodiversity intactness'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohibc/src/templates/ohibc_hdr1.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(raster)
library(data.table)

source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')
  ### includes library(tidyverse); library(stringr); 
  ### dir_M points to ohi directory on Mazu; dir_O points to home dir on Mazu

dir_git <- '~/github/spp_risk_dists'

### goal specific folders and info
dir_setup <- file.path(dir_git, 'data_setup')
dir_data  <- file.path(dir_git, 'data')
dir_o_anx <- file.path(dir_O, 'git-annex/spp_risk_dists')

source(file.path(dir_git, 'data_setup/api_fxns.R'))

```

# Summary

Compare MPAs to biodiversity intactness: 

# Methods

<!--
## Examine variance against mean

Examine correlation between risk and variance?  A cell with medium risk and high variance indicates balance of low-risk species with some at risk species.  A cell with medium risk and low variance indicates a more systemically at-risk cell.-->

``` {r var vs mean, eval = FALSE}

mean_rast   <- raster(file.path(dir_git, 'output', 'mean_risk_raster_gp.tif'))
var_rast    <- raster(file.path(dir_git, 'output', 'var_risk_raster_gp.tif'))
n_spp_rast  <- raster(file.path(dir_git, 'output', 'n_spp_risk_raster_gp.tif'))


mean_var_df <- data.frame(mean_risk = values(mean_rast),
                          var_risk  = values(var_rast),
                          n_spp     = values(n_spp_rast)) %>%
  filter(!is.na(mean_risk)) %>%
  filter(n_spp >= 5)

mean_var_sample <- sample_n(mean_var_df, 100000)

ggplot(mean_var_sample, aes(x = mean_risk, y = var_risk)) +
  ggtheme_plot() +
  geom_point(aes(color = n_spp), alpha = .1) +
  scale_color_gradientn(colors = c('grey90', 'lightblue', 'blue1', 'blue3', 'blue4', 'purple4'),
                        values = c(0, .01, .05, .1, .25, 1))

coef_var_sample <- mean_var_df %>%
  filter(mean_risk != 0) %>%
  filter(n_spp > 5) %>%
  sample_n(100000) %>%
  mutate(coef_var = sqrt(var_risk) / mean_risk,
         inv_cv   = 1 / coef_var)
 
ggplot(coef_var_sample, aes(x = mean_risk, y = n_spp)) +
  ggtheme_plot() +
  geom_point(aes(color = n_spp), alpha = .1) +
  scale_color_gradientn(colors = c('grey90', 'lightblue', 'blue1', 'blue3', 'blue4', 'purple4'),
                        values = c(0, .01, .05, .1, .25, 1))

```

<!--Inverse coefficient of variation, i.e. mean risk / standard deviation of risk, will tend to be higher for high means that correspond to low variance.  Low variance occurs when all species within a region are similar in risk; this indicates a system-wide moderate risk.  High variance occurs when species risks differ from the central tendency; this indicates a system with low-risk species and high risk species, most likely skewed toward low risk as long as mean is closer to 0 than 1.  High ICV probably is a good candidate for MPAs.-->

## Compare MPA coverage to biodiversity risk

MPA coverage will be compared to non-MPA areas within EEZs.  The .csv codes MPAs as their numeric IUCN category (i.e. Ia = Ib = 1, II = 2, ..., VI = 6); areas that are specifically designated as no-take (that are not already in categories Ia, Ib, and II) are coded as -1; areas that are designated in the WDPA database but do not fall into one of these categories are coded as 8.

``` {r set up mpa and values dataframe}

mean_rast     <- raster(file.path(dir_git, 'output', 'mean_risk_raster_gp.tif'))
var_rast      <- raster(file.path(dir_git, 'output', 'var_risk_raster_gp.tif'))
n_spp_rast    <- raster(file.path(dir_git, 'output', 'n_spp_risk_raster_gp.tif'))
n_threat_rast <- raster(file.path(dir_git, 'output', 'n_threat_raster_gp.tif'))
trend_rast    <- raster(file.path(dir_git, 'output', 'trend_raster_gp.tif'))
eez_rast      <- raster(file.path(dir_git, 'spatial', 'eez_rast_gp.tif'))
mpa_df        <- read_csv(file.path(dir_git, 'spatial', 'wdpa_mpa_area_gp.csv'))

# tmp_rast <- calc(eez_rast, fun = function(x) x %in% c(1:212, 214:255))
### excludes high seas and antarctica

risk_df <- data.frame(cell_id   = 1:length(values(mean_rast)),
                      eez       = values(eez_rast),
                      mean_risk = values(mean_rast),
                      var_risk  = values(var_rast),
                      n_spp     = values(n_spp_rast),
                      n_threat  = values(n_threat_rast),
                      mean_trend = values(trend_rast)) %>%
  mutate(icv = mean_risk / sqrt(var_risk)) %>%
  filter(eez %in% c(1:212, 214:254)) %>% 
    ### 213 is Antarctica, 255 is disputed territories, >255 is high seas
  filter(!is.na(mean_risk))

risk_mpa_df <- risk_df %>%
  left_join(mpa_df, by = 'cell_id') %>%
  mutate(mpa_pct = ifelse(is.na(mpa_pct), 0, mpa_pct),
         wdpa_category = ifelse(is.na(wdpa_category), 0, wdpa_category)) %>%
  left_join(read_csv(file.path(dir_git, 'spatial/rgn_names.csv')), by = c('eez' = 'rgn_id')) %>%
  mutate(r1_label = str_replace(r1_label, 'Latin Am.+', 'Latin America'))

```

Consider a cell "protected" if the appropriate level of protection covers at least half the cell.

``` {r, eval = TRUE}

global_no_take <- risk_mpa_df %>%
  mutate(lbl = 'Global') %>%
  select(cell_id, wdpa_category, mean_risk, mpa_pct, lbl) %>%
  mutate(mpa = wdpa_category <= 2 & mpa_pct > .5,
         protection = 'Ia, Ib, II, no-take')

rgn_no_take <- risk_mpa_df %>%
  mutate(lbl = r1_label) %>%
  arrange(lbl) %>%
  select(cell_id, wdpa_category, mean_risk, mpa_pct, lbl) %>%
  mutate(mpa = wdpa_category <= 2 & mpa_pct > .5,
         protection = 'Ia, Ib, II, no-take')

no_take_df <- bind_rows(global_no_take, rgn_no_take) %>%
  mutate(lbl = forcats::fct_inorder(lbl))
# no_take_df$lbl %>% levels()

no_take_means <- no_take_df %>%
  group_by(mpa, lbl) %>%
  summarize(mu = mean(mean_risk), sd = sd(mean_risk, na.rm = TRUE), se = sd / sqrt(n()))
  

```



``` {r set up overall rr-weighted dataframe}

mean_rr_rast      <- raster(file.path(dir_git, 'output', 'mean_rr_risk_raster_gp.tif'))
var_rr_rast       <- raster(file.path(dir_git, 'output', 'var_rr_risk_raster_gp.tif'))
sr_rr_rast        <- raster(file.path(dir_git, 'output', 'sr_rr_risk_raster_gp.tif'))
sr_rr_threat_rast <- raster(file.path(dir_git, 'output', 'sr_rr_threat_raster_gp.tif'))
trend_rr_rast     <- raster(file.path(dir_git, 'output', 'trend_rr_raster_gp.tif'))

risk_rr_df <- data.frame(cell_id   = 1:length(values(mean_rast)),
                         eez       = values(eez_rast),
                         mean_risk  = values(mean_rr_rast),
                         var_risk      = values(var_rr_rast),
                         sr_rr_risk    = values(sr_rr_rast),
                         sr_rr_threat  = values(sr_rr_threat_rast),
                         mean_rr_trend = values(trend_rr_rast)) %>%
  filter(!is.na(mean_risk)) %>%
  filter(eez %in% c(1:212, 214:254)) %>% ### 255 is disputed territories
  left_join(read_csv(file.path(dir_git, 'spatial/rgn_names.csv')), by = c('eez' = 'rgn_id'))

risk_mpa_rr_df <- risk_rr_df %>%
  left_join(mpa_df, by = 'cell_id') %>%
  mutate(mpa_pct = ifelse(is.na(mpa_pct), 0, mpa_pct),
         wdpa_category = ifelse(is.na(wdpa_category), 0, wdpa_category)) %>%
  mutate(r1_label = str_replace(r1_label, 'Latin Am.+', 'Latin America'))

```

``` {r rr-weighted protection level dataframes}
global_no_take_rr <- risk_mpa_rr_df %>%
  mutate(lbl = 'Global') %>%
  select(cell_id, wdpa_category, mean_risk, mpa_pct, lbl) %>%
  mutate(mpa = wdpa_category <= 2 & mpa_pct > .5,
         protection = 'Ia, Ib, II, no-take')

rgn_no_take_rr <- risk_mpa_rr_df %>%
  mutate(lbl = r1_label) %>%
  arrange(lbl) %>%
  select(cell_id, wdpa_category, mean_risk, mpa_pct, lbl) %>%
  mutate(mpa = wdpa_category <= 2 & mpa_pct > .5,
         protection = 'Ia, Ib, II, no-take')

no_take_rr_df <- bind_rows(global_no_take_rr, rgn_no_take_rr) %>%
  mutate(lbl = forcats::fct_inorder(lbl))
# no_take_df$lbl %>% levels()

no_take_rr_means <- no_take_rr_df %>%
  group_by(mpa, lbl) %>%
  summarize(mu = mean(mean_risk), sd = sd(mean_risk, na.rm = TRUE), se = sd / sqrt(n()))
  
```

There seem to be two philosophies of MPA priority: place MPAs in impacted areas to reduce pressure and prevent further degradatation, or place MPAs to protect pristine places to maintain their pristine state.  This suggests that we may see a bimodal distribution of mean risk within MPA cells relative to the distribution of mean risk in general.

### Mean risk

``` {r plot_function}

plot_dist <- function(df, means_df, var_name, x_lims = c(0, .75)) {
  x <- ggplot(df, aes(x = mean_risk, fill = mpa, color = mpa)) +
    ggtheme_plot() +
    theme(strip.text.y = element_text(angle = 0, hjust = 0),
          axis.text.y  = element_blank(),
          axis.title   = element_blank(),
          panel.grid.major.y = element_blank()) +
    geom_density(alpha = .5, size = .25) +
    geom_vline(data = means_df, 
               aes(xintercept = mu, linetype = mpa),
               alpha = .8) +
    scale_x_continuous(expand = c(0, 0), 
                       limits = x_lims,
                       labels = c('LC', 'NT', 'VU', 'EN', 'CR', 'EX'),
                       breaks = c( 0.0,  0.2,  0.4,  0.6,  0.8,  1.0)) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_fill_manual(values  = c('red4', 'cadetblue2')) +
    scale_color_manual(values = c('red4', 'cadetblue3')) +
    facet_grid( lbl ~ ., scales = 'free_y') +
    labs(fill = var_name, color = var_name, linetype = var_name)
  
  return(x)
}

```

``` {r}

mpa_plot_unweighted <- plot_dist(no_take_df, no_take_means, 'No take')

mpa_plot_rrweighted <- plot_dist(no_take_rr_df, no_take_rr_means, 'No take')

plot_combined <- cowplot::plot_grid(mpa_plot_unweighted +
                                      theme(strip.text.y = element_blank(),
                                            legend.justification = c(1, 1), 
                                            legend.position = c(1, 1)), 
                                    mpa_plot_rrweighted +
                                      theme(legend.position = 'none'), 
                                    labels = c('A', 'B'),
                                    rel_widths = c(3, 4))

ggsave(file.path(dir_git, 'ms_figures/fig3_bd_risk_vs_mpas.png'),
       height = 4, width = 6, dpi = 300, units = 'in')
```

