---
title: 'Biodiversity n and var two-panel figures'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohibc/src/templates/ohibc_hdr1.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(raster)
library(data.table)
library(sf)


source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')  ###
  ### includes library(tidyverse); library(stringr); dir_M points to ohi directory

dir_git <- '~/github/spp_risk_dists'

### goal specific folders and info
dir_data  <- file.path(dir_git, 'data')
dir_o_anx <- file.path(dir_O, 'git-annex/spp_risk_dists')

source(file.path(dir_git, 'setup/common_fxns.R'))

### Gall-Peters doesn't have an EPSG?
gp_proj4 <- '+proj=cea +lon_0=0 +lat_ts=45 +x_0=0 +y_0=0 +ellps=WGS84 +units=m +no_defs'

```

# Summary

Create a set of maps of the distribution of biodiversity intactness - all species assessed and mapped by IUCN.  These maps are generated at 10 km^2^ resolution in a Gall-Peters projection.  These maps will be generated using all comprehensively-assessed species:

* S1: Number of species for mean/var calculations
* S3: Variance of risk

# Data Sources

IUCN Red List spatial data download
IUCN Red List API
Gina Ralph (IUCN)

# Methods

## Spatial distribution of N spp: comprehensively assessed spp

From the 1a and 1b biodiversity risk map scripts, gather the rasters for the various maps:

* N_spp risk (unweighted and range-rarity-weighted)
* Var risk (unweighted and range-rarity-weighted)

``` {r get_rasts}

nspp_un_rast <- raster(file.path(dir_git, 'output', 'n_spp_risk_raster_comp.tif'))
nspp_rr_rast <- raster(file.path(dir_git, 'output', 'sr_rr_risk_raster_comp.tif'))
var_un_rast  <- raster(file.path(dir_git, 'output', 'var_risk_raster_comp.tif'))
var_rr_rast  <- raster(file.path(dir_git, 'output', 'var_rr_risk_raster_comp.tif'))

```

### And now, the maps

The four-panel map chunks create two maps, with two density plots as scales; first map is for uniform-weighting, second is for range-rarity weighting.  This is slightly different from that used for Fig 1 in the main text; that put uniform and rr vertically, while this does it side by side.


``` {r fig S1 a b n_spp maps}

### Log-transform nspp rasters for more interesting visual scaling
nspp_un_log <- log10(nspp_un_rast) # range 0, 3.146748
nspp_rr_log <- log10(nspp_rr_rast) # -8.23878, -0.9510185
### aggregate rasters for faster testing:
# nspp_un_rast <- nspp_un_rast %>% aggregate(10)
# nspp_rr_rast <- nspp_rr_rast %>% aggregate(10)

nspp_cols <- c(viridis::viridis(4, direction = -1))

land_poly <- sf::read_sf(file.path(dir_git, 'spatial/ne_10m_land/ne_10m_land.shp')) %>%
  st_transform(gp_proj4)

map_df <- data.frame(val_1 = values(nspp_un_log),
                     val_2 = values(nspp_rr_log)) %>%
  cbind(coordinates(map1_rast)) %>% 
  filter(!is.na(val_1))

map1 <- ggplot(map_df) +
  geom_raster(aes(x, y, fill = val_1), show.legend = FALSE) +
  geom_sf(data = land_poly, aes(geometry = geometry), 
          fill = 'grey96', color = 'grey40', size = .10) +
  ggtheme_map() +
  theme(plot.margin = unit(c(.2, 0, .1, .5), units = 'cm')) +
  coord_sf(datum = NA) + ### ditch graticules
  scale_fill_gradientn(colors = nspp_cols) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0))

map2 <- ggplot(map_df) +
  geom_raster(aes(x, y, fill = val_2), show.legend = FALSE) +
  geom_sf(data = land_poly, aes(geometry = geometry), 
          fill = 'grey96', color = 'grey40', size = .10) +
  ggtheme_map() +
  theme(plot.margin = unit(c(.1, 0, .2, .5), units = 'cm')) +
  coord_sf(datum = NA) + ### ditch graticules
  scale_fill_gradientn(colors = nspp_cols) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0))

###################

### set up a dataframe of values to craft a color bar using geom_segment
colorbar_df1 <- data.frame(x = seq(0, max(map_df$val_1), length.out = 1000), y = -1)

lbls1 = c(1, 10, 100, 1000); brks1 = c(0, 1, 2, 3)
dens1 <- ggplot(map_df) +
  ggtheme_plot() +
  geom_vline(xintercept = mean(map_df$val_1, na.rm = TRUE)) +
  geom_segment(data = colorbar_df1, 
               aes(x = x, xend = x, color = x), 
               y = 0, yend = -10, size = 1,
               show.legend = FALSE) +
  geom_density(aes(x = val_1, ..scaled..), adjust = 5,
               alpha = .3, size = .25, fill = 'grey30') +
  scale_color_gradientn(colors = nspp_cols, 
                        labels = lbls1, breaks = brks1) +
  theme(axis.text.x  = element_blank(), 
        axis.title = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.margin = unit(c(1, 0, 1, 0), units = 'cm')) +
  scale_x_continuous(labels = lbls1, breaks = brks1,
                     expand = c(0, 0)) +
  scale_y_continuous(limits = c(-.4, 1)) +
  coord_flip()
  
colorbar_df2 <- data.frame(x = seq(min(map_df$val_2), 
                                   max(map_df$val_2), 
                                   length.out = 1000), 
                           y = -1)
lbls2 = c('1e-8', '1e-6', '1e-4', '1e-2', '1'); brks2 = c(-8, -6, -4, -2, 0)

dens2 <- ggplot(map_df) +
  ggtheme_plot() +
  geom_vline(xintercept = mean(map_df$val_2, na.rm = TRUE)) +
  geom_segment(data = colorbar_df2, 
               aes(x = x, xend = x, color = x), 
               y = 0, yend = -10, size = 1,
               show.legend = FALSE) +
  geom_density(aes(x = val_2, ..scaled..), 
               adjust = 5,
               alpha = .3, size = .25, fill = 'grey30') +
  scale_color_gradientn(colors = nspp_cols, 
                        labels = lbls2, breaks = brks2) +
  theme(axis.text.x  = element_blank(), 
        axis.title = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.margin = unit(c(1, 0, 1, 0), units = 'cm')) +
  scale_x_continuous(labels = lbls2, breaks = brks2,
                     expand = c(0, 0)) +
  scale_y_continuous(limits = c(-.4, 1)) +
  coord_flip()

panel_left <- cowplot::plot_grid(map1, dens1, 
                                axis = 'b',
                                rel_widths = c(5, 1))
panel_right <- cowplot::plot_grid(map2, dens2, 
                                axis = 'b',
                                rel_widths = c(5, 1))

four_panel <- cowplot::plot_grid(panel_left, panel_right, 
                                 labels = c('A', 'B'),
                                 nrow = 1, ncol = 2, align = 'h')

figS1_path <- file.path(dir_git, 'ms_figures/fig_S1_nspp_maps.png')
ggsave(plot = four_panel,
       filename = figS1_path,
       width = 10, height = 2.5, units = 'in', dpi = 300)


```

``` {r fig S3 a b var maps}

var_cols <- c(viridis::viridis(4, direction = -1))

land_poly <- sf::read_sf(file.path(dir_git, 'spatial/ne_10m_land/ne_10m_land.shp')) %>%
  st_transform(gp_proj4)

map_df <- data.frame(val_1 = values(var_un_rast),
                     val_2 = values(var_rr_rast)) %>%
  cbind(coordinates(map1_rast)) %>% 
  filter(!is.na(val_1))
### unweighted max = .18
### rr weighted max = .499

map1 <- ggplot(map_df) +
  geom_raster(aes(x, y, fill = val_1), show.legend = FALSE) +
  geom_sf(data = land_poly, aes(geometry = geometry), 
          fill = 'grey96', color = 'grey40', size = .10) +
  ggtheme_map() +
  theme(plot.margin = unit(c(.2, 0, .1, .5), units = 'cm')) +
  coord_sf(datum = NA) + ### ditch graticules
  scale_fill_gradientn(colors = var_cols) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0))

map2 <- ggplot(map_df) +
  geom_raster(aes(x, y, fill = val_2), show.legend = FALSE) +
  geom_sf(data = land_poly, aes(geometry = geometry), 
          fill = 'grey96', color = 'grey40', size = .10) +
  ggtheme_map() +
  theme(plot.margin = unit(c(.1, 0, .2, .5), units = 'cm')) +
  coord_sf(datum = NA) + ### ditch graticules
  scale_fill_gradientn(colors = var_cols) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0))

###################

### set up a dataframe of values to craft a color bar using geom_segment
colorbar_df <- data.frame(x = seq(0, .5, length.out = 1000), y = -1)

lbls = c(0, .1, .2, .3, .4); brks = c(0, .1, .2, .3, .4)

dens1 <- ggplot(map_df) +
  ggtheme_plot() +
  geom_vline(xintercept = mean(map_df$val_1, na.rm = TRUE)) +
  geom_segment(data = colorbar_df, 
               aes(x = x, xend = x, color = x), 
               y = 0, yend = -10, size = 1,
               show.legend = FALSE) +
  geom_density(aes(x = val_1, ..scaled..), adjust = 5,
               alpha = .3, size = .25, fill = 'grey30') +
  scale_color_gradientn(colors = var_cols, 
                        labels = lbls, breaks = brks) +
  theme(axis.text.x  = element_blank(), 
        axis.title = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.margin = unit(c(1, 0, 1, 0), units = 'cm')) +
  scale_x_continuous(labels = lbls, breaks = brks,
                     expand = c(0, 0)) +
  scale_y_continuous(limits = c(-.4, 1)) +
  coord_flip()
  
dens2 <- ggplot(map_df) +
  ggtheme_plot() +
  geom_vline(xintercept = mean(map_df$val_2, na.rm = TRUE)) +
  geom_segment(data = colorbar_df, 
               aes(x = x, xend = x, color = x), 
               y = 0, yend = -10, size = 1,
               show.legend = FALSE) +
  geom_density(aes(x = val_2, ..scaled..), 
               adjust = 5,
               alpha = .3, size = .25, fill = 'grey30') +
  scale_color_gradientn(colors = var_cols, 
                        labels = lbls, breaks = brks) +
  theme(axis.text.x  = element_blank(), 
        axis.title = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.margin = unit(c(1, 0, 1, 0), units = 'cm')) +
  scale_x_continuous(labels = lbls, breaks = brks,
                     expand = c(0, 0)) +
  scale_y_continuous(limits = c(-.4, 1)) +
  coord_flip()

panel_left <- cowplot::plot_grid(map1, dens1, 
                                axis = 'b',
                                rel_widths = c(5, 1))
panel_right <- cowplot::plot_grid(map2, dens2, 
                                axis = 'b',
                                rel_widths = c(5, 1))

four_panel <- cowplot::plot_grid(panel_left, panel_right, 
                                 labels = c('A', 'B'),
                                 nrow = 1, ncol = 2, align = 'h')

figS3_path <- file.path(dir_git, 'ms_figures/fig_S3_var_maps.png')
ggsave(plot = four_panel,
       filename = figS3_path,
       width = 10, height = 2.5, units = 'in', dpi = 300)


```
