---
title: 'Process IUCN spp shapes'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/src/templates/ohara_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(raster)
library(rgeos)
source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')

library(sf)

dir_git <- '~/github/spp_risk_dists'
dir_o_anx <- file.path(dir_O, 'git-annex/spp_risk_dists')

### project specific folders and info
source(file.path(dir_git, 'data_setup/common_fxns.R'))

dir_shp <- file.path(dir_M, 'git-annex/globalprep/_raw_data/iucn_spp/d2017-3')
dir_bli <- file.path(dir_M, 'git-annex/globalprep/_raw_data/birdlife_intl/d2018')
dir_shp2 <- file.path(dir_M, 'git-annex/globalprep/_raw_data/iucn_spp/d2018-1')
  ### These are shapefiles directly from IUCN as individual species map files
  ### (in the unzipped folder).
  ### in this folder are shapefiles at a taxonomic level.

### Gall-Peters doesn't have an EPSG?
gp_proj4 <- '+proj=cea +lon_0=0 +lat_ts=45 +x_0=0 +y_0=0 +ellps=WGS84 +units=m +no_defs'

```

# Summary

Using a set of IUCN species range maps, rasterize each species to 0.1° lat long raster using `fasterize`.  Use `presence` field from shapefile.

* Subpopulation polygons must be identified and rasterized separately from the parent polygon; this must be done by sciname and subpop fields since the polygon IDs are based upon the parent ID.
* Regional assessments need not be determined at this stage - the ID numbers match the global ID numbers (including subpops).

# Data source

IUCN Red List: Spatial Data Download
IUCN: Gina Ralph direct communication

# Methods

## Generate species map list with subpops

### Create subpopulation lookup to match maps with IUCN SIDs for subpops

From available shapefiles, reclassify subpopulation polygons into subpop ID numbers for matching to risk and trend. 

``` {r convert BLI to shp-dbf-etc}

### convert BirdLife Int'l .gdb to shapefile, after filtering to species in the "marine" list.
### Here we also standardize the column names to match the IUCN column names.

bli_marine_shp <- file.path(dir_bli, 'bli_marine_v2017.shp')

if(!file.exists(bli_marine_shp)) {
  
  bli_sf <- st_read(file.path(dir_bli, 'BOTW.gdb'))
  
  marine_spp <- read_csv(file.path(dir_git, sprintf('data/spp_marine_from_api_%s.csv', api_version)))
  
  bli_marine_sf <- bli_sf %>%
    clean_df_names() %>%
    filter(sisid %in% marine_spp$iucn_sid)
  
  st_write(bli_marine_sf, bli_marine_shp)
  
  ### clean column names
  bli_marine_dbf <- str_replace(bli_marine_shp, '\\.shp$', '\\.dbf')
  
  att_table <- foreign::read.dbf(bli_marine_dbf) %>%
    setNames(names(bli_marine_sf)[1:(ncol(bli_marine_sf) - 1)]) %>%
    rename(iucn_sid = sisid, binomial = sciname, year = date, tax_comm = tax_com) %>%
    mutate(year = as.integer(as.character(year)))
    #  [1] "sisid"        "sciname"      "date"         "source"       "presence"     "origin"       "seasonal"    
    #  [8] "data_sens"    "sens_comm"    "compiler"     "tax_com"      "dist_com"     "reviewers"    "citation"    
    # [15] "version"      "shape_length" "shape_area"  
  ### from IUCN shapefiles:
    #  [1] "id_no"      "binomial"   "presence"   "origin"     "seasonal"   "compiler"   "year"       "citation"  
    #  [9] "source"     "dist_comm"  "island"     "subspecies" "subpop"     "legend"     "tax_comm"   "kingdom"   
    # [17] "phylum"     "class"      "order_"     "family"     "genus"      "code"       "marine"     "terrestial"
    # [25] "freshwater" "shape_Leng" "shape_Area"
  foreign::write.dbf(att_table, bli_marine_dbf)
}

```

Note that there appear to be no subpopulation polygons within the BirdLife International map set; there is no "subpop" column in that attribute table.  It appears that the individual files may not have "subpop" column either.

``` {r create_subpop_lookup}

iucn_dbfs <- list.files(dir_shp, pattern = '\\.dbf$', full.names = TRUE)
iucn2_dbfs <- list.files(dir_shp2, pattern = '\\.dbf$', recursive = FALSE,
                         full.names = TRUE)
bli_dbfs  <- list.files(dir_bli, pattern = '\\.dbf$', full.names = TRUE)
shps_dbfs <- c(iucn_dbfs, bli_dbfs, iucn2_dbfs)

### This is all "marine" species by habitat:
marine_spp <- read_csv(file.path(dir_git, sprintf('data/spp_marine_from_api_%s.csv', api_version)))

### This is *all* spp from the API, filtered down to those in marine_spp: 
api_spp <- read_csv(file.path(dir_o_anx, 'iucn', 
                              sprintf('spp_info_from_api_%s.csv', api_version))) %>%
  filter(iucn_sid %in% marine_spp$iucn_sid)

### Define helper function to cleanly get the info from DBF files,
### including colname checks
get_dbf <- function(x) { ### x <- shps_dbfs[1]
  spp_dbf_info <- foreign::read.dbf(x, as.is = TRUE) %>%
    clean_df_names()
  
  ### Individual files call the id number 'iucn_sid'; bli has been fixed to
  ### 'iucn_sid' as well; Red List Data Download files are 'id_no'... So:
  ### fix that here!
  names(spp_dbf_info)[names(spp_dbf_info) == 'id_no'] <- 'iucn_sid'
  
  ### Some files (e.g. bli and individual files) don't have 'subpop':
  ### add it in as NAs.
  if(!'subpop' %in% names(spp_dbf_info)) spp_dbf_info$subpop <- NA
  
  ### some files don't have 'presence' field: if not present or NA,
  ### set to presence = 1.  Some have presence = 0?
  if(!'presence' %in% names(spp_dbf_info)) spp_dbf_info$presence <- 1
  
  spp_dbf_info <- spp_dbf_info %>%
    mutate(presence = ifelse(presence == 0, 1, presence))
  
  return(spp_dbf_info)
}

shps_df <- lapply(shps_dbfs, get_dbf) %>%
  setNames(basename(shps_dbfs)) %>%
  bind_rows(.id = 'dbf_file') %>%
  select(dbf_file, iucn_sid, sciname = binomial, presence, subpop) %>%
  filter(iucn_sid %in% marine_spp$iucn_sid)

shp_subpops <- shps_df %>%
  select(shp_iucn_sid = iucn_sid, sciname, shp_subpop = subpop) %>%
  filter(!is.na(shp_subpop)) %>%
  distinct() %>%
  mutate(shp_subpop_clean = str_replace(tolower(shp_subpop), ' subpopulation| ocean', ''),
         shp_subpop_clean = str_replace_all(shp_subpop_clean, '[^a-z]', ' ') %>% str_trim,
         shp_subpop_clean = str_replace(shp_subpop_clean, 'noth', 'north'), 
           ### fix typo
         shp_subpop_clean = ifelse(str_detect(shp_subpop_clean, 'svalbard barents sea'),
                                   'east greenland svalbard barents sea', shp_subpop_clean))
           ### fix mismatch between 2018-1 api subpop and 2017-3 shp subpop

api_subpops <- api_spp %>%
  select(api_iucn_sid = iucn_sid, sciname, api_subpop = population) %>%
  filter(!is.na(api_subpop)) %>%
  filter(sciname %in% shp_subpops$sciname) %>%
  distinct() %>%
  mutate(api_subpop_clean = str_replace_all(tolower(api_subpop), ' subpopulation| ocean', ''),
         api_subpop_clean = str_replace_all(api_subpop_clean, '[^a-z]', ' ') %>% str_trim)

subpops_match_raw <- shp_subpops %>%
  full_join(api_subpops, by = 'sciname') %>%
  group_by(api_iucn_sid) %>%
  mutate(subpop_match = str_detect(shp_subpop_clean, api_subpop_clean),
         n_match = sum(subpop_match)) %>%
  filter(subpop_match | sum(subpop_match) == 0) %>%
  ungroup()

caretta_subpops <- subpops_match_raw %>%
  filter(shp_iucn_sid == 3897) %>%
  mutate(api_subpop_single = str_split(api_subpop_clean, ' ')) %>%
  unnest(api_subpop_single) %>%
  group_by(api_subpop_clean, shp_subpop_clean) %>%
  mutate(match = str_detect(shp_subpop_clean, api_subpop_single),
         n_match = sum(match),
         n_words = n()) %>%
  filter(sum(match) == n()) %>%
  ungroup() %>%
  select(shp_iucn_sid, sciname, shp_subpop, api_iucn_sid, api_subpop) %>%
  distinct()

subpops_match <- subpops_match_raw %>%
  filter(shp_iucn_sid != 3897 & subpop_match == TRUE) %>%
  select(shp_iucn_sid, sciname, shp_subpop, api_iucn_sid, api_subpop) %>%
  bind_rows(caretta_subpops)

### Check for missed matches:

# api_subpops$api_iucn_sid[!api_subpops$api_iucn_sid %in% subpops_match$api_iucn_sid]
### API subpop, not matched: 16369383 Tursiops truncatus Mediterranean subpopulation
### OK - no polygon for this subpopulation.  Make sure it gets picked up by regional assessment!
### API subpop, not matched: 2472 Balaena mysticetus East Greenland-Svalbard-Barents Sea subpopulation
### NOOO! This is called something slightly different between shp and api.
### * Code above is fixed for this special case.  Note, if/when 2018-1 shapefiles are
###   added in, this code may need to come out or be adjusted.

# shp_subpops$shp_subpop[!shp_subpops$shp_subpop %in% subpops_match$shp_subpop]
### Shapefile subpop, not matched: 16285718 Chelonia mydas Hawaiian subpopulation
### OK - the shapefile ID matches the API ID so this will match automatically

write_csv(subpops_match, 
          file.path(dir_git, sprintf('data_setup/int/subpops_match_api_to_shp_%s.csv', 
                                     api_version)))

```

### get info on comprehensive assessments

Not all maps are for species that are included in comprehensively assessed taxonomic groups.  Get info on comprehensive assessment status for species from the API.

``` {r get comp assess status}

comp_gps_url <- 'http://apiv3.iucnredlist.org/api/v3/comp-group/list?token=%s'
comp_gps <- fromJSON(sprintf(comp_gps_url, api_key)) %>%
  .$result

comp_spp_url <- 'http://apiv3.iucnredlist.org/api/v3/comp-group/getspecies/%s?token=%s'
comp_spp <- mc_get_from_api(comp_spp_url, comp_gps$group_name, api_key)
```

### Join map list to subpop corrected list; remove duplicated spp IDs

From the species range map shapefiles, pull map info from the .dbf files.  For species listed in multiple files (e.g. sea snakes and reptiles), remove duplicates.  For species with polygons differentiated by subpopulation, adjust the iucn_sid to match the subpop info.

``` {r generate map list from dbfs}

iucn_dbfs  <- list.files(dir_shp, pattern = '\\.dbf$', full.names = TRUE)
iucn2_dbfs <- list.files(dir_shp2, pattern = '\\.dbf$', recursive = FALSE,
                         full.names = TRUE)
bli_dbfs   <- list.files(dir_bli, pattern = '\\.dbf$', full.names = TRUE)

map_files  <- c(iucn_dbfs, bli_dbfs, iucn2_dbfs)


# spp_group_names <- map_files %>%
#   basename() %>%
#   str_replace_all('\\.dbf$|_PART_.+', '') %>%
#   unique()

map_info_list <- lapply(map_files, get_dbf) %>%
  setNames(map_files)

map_info_raw <- bind_rows(map_info_list, .id = 'dbf_file') %>%
  select(shp_iucn_sid = iucn_sid, 
         sciname = binomial, 
         presence, 
         subpop, 
         dbf_file)
### Notes on presence, origin, seasonal fields:
### * presence = 5 is extinct; 4 = probably extinct; others are extant-ish or 
###   uncertain. We will drop field and include all polygons for now.
### * origin is native, introduced, etc.  We will drop this field and not
###   worry about origin.
### * seasonal is breeding/non-breeding/passage.  We will drop this field
###   and not worry about seasonality.

# subspp <- map_info %>%
#   filter(!is.na(subspecies)) %>%
#   select(id_no, code) %>%
#   distinct()
### No subspecies vary by IUCN code; drop field for simplicity
### and include all at species level

map_info_sans_dupes <- map_info_raw %>%
  group_by(shp_iucn_sid) %>%
  filter((length(unique(dbf_file)) == 1) | 
           (!str_detect(dbf_file, 'MARINEFISH|REPTILE|CORALS_PART_2_1|MANGROVES'))) %>%
  filter(!(shp_iucn_sid == 196026 & str_detect(dbf_file, 'SEASNAKE'))) %>%
  ungroup() %>%
  distinct() # %>%
  # group_by(shp_iucn_sid) %>%
  # mutate(n_dbfs = length(unique(dbf_file))) %>%
  # ungroup()

subpops_match <- read_csv(file.path(dir_git, 'data_setup/int',
                                    sprintf('subpops_match_api_to_shp_%s.csv', 
                                            api_version)),
                          col_types = 'dccdc')

map_info_add_subpops <- map_info_sans_dupes %>%
  full_join(subpops_match, by = c('shp_iucn_sid', 'sciname', 'subpop' = 'shp_subpop')) %>%
  mutate(iucn_sid = ifelse(is.na(api_iucn_sid), shp_iucn_sid, api_iucn_sid)) %>%
  select(shp_iucn_sid, iucn_sid, sciname, presence, subpop, dbf_file) %>%
  distinct()

### filter to marine habitat species (to drop terrestrial reptiles e.g.)
marine_spp_ids <- read_csv(file.path(dir_git, 'data',
                                     sprintf('spp_marine_from_api_%s.csv', 
                                             api_version)),
                           col_types = 'dcc') %>%
  select(iucn_sid, max_depth)

marine_map_info <- map_info_add_subpops %>%
  inner_join(marine_spp_ids, by = 'iucn_sid') %>%
  mutate(comp_assessed = iucn_sid %in% comp_spp$taxonid)

write_csv(marine_map_info, 
          file.path(dir_git, 'data',
                    sprintf('spp_marine_maps_%s.csv', api_version)))

```
