---
title: 'Aggregate IUCN spp ranges'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/src/templates/ohara_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(raster)
source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')

library(sf)
library(data.table)

dir_git <- '~/github/spp_risk_dists'
dir_o_anx <- file.path(dir_O, 'git-annex/spp_risk_dists')
dir_setup <- file.path(dir_git, '_setup')

source(file.path(dir_setup, 'common_fxns.R'))

### goal specific folders and info

dir_data <- file.path(dir_git, '_data')

```

# Summary

Using individual species range data (csv outputs from `5_process_spp_shps.Rmd`), collect into species groups.  In each species group, process range and risk information to get mean risk, variance of risk, species richness, and threatened species count per cell.  These will then be analyzed in the context of species range rarity weighting as well in a sister script.

# Methods

## Generate species group maps - equal weighting per cell

From species maps data.frame, identify the IDs of species within each taxon.  For each taxon:

* Collect all taxon species into a single data.frame.
* Join species extinction risk and trend data.frame to species-cell data.frame, including regional assessments.
* By cell, calculate mean extinction risk, variance of extinction risk, trend, threatened species count, n_spp, n_spp_trend
    * When aggregating further, use n_spp to weight values.
    
__NOTE:__ Because the summary files are likely to be very large for globe-spanning taxa (e.g. MARINE_MAMMALS is 275 MB), save these outputs outside of GitHub.

``` {r aggregate species maps to taxa}

### Read in lots of data
spp_maps <- read_csv(file.path(dir_data, 
                               sprintf('spp_marine_maps_%s.csv', api_version)),
                     col_types = 'ddcicccl')

spp_risk <- read_csv(file.path(dir_data, 
                               sprintf('iucn_risk_current_%s.csv', api_version)),
                     col_types = 'dccicccdc')

spp_risk_rgn <- read_csv(file.path(dir_data, 
                                   sprintf('iucn_risk_rgn_current_%s.csv', api_version)),
                         col_types = 'dcccdc')

spp_trend <- read_csv(file.path(dir_data, sprintf('iucn_trend_by_spp_%s.csv', api_version)),
                      col_types = 'dccdc') %>%
  select(iucn_sid, trend_score, iucn_rgn)

### make a dataframe of species risk, trend, and regional risk
spp_risk_trend <- spp_risk %>%
  mutate(iucn_rgn = 'global') %>%
  bind_rows(spp_risk_rgn) %>%
  select(iucn_sid, iucn_rgn, cat_score) %>%
  left_join(spp_trend, by = c('iucn_sid', 'iucn_rgn'))

### raster for cell IDs
rast_cell_ids <- raster(file.path(dir_git, '_spatial', 'cell_id_rast.tif'))

### Make a dataframe of cell ID to MEOW for regional assessments... also 
### make a lookup of MEOW to region
meow_rgns_rast <- raster::raster(file.path(dir_git, '_spatial', 'meow_rast.tif'))
meow_cells_all <- data.frame(cell_id = values(rast_cell_ids),
                             meow_id = values(meow_rgns_rast)) %>%
  filter(!is.na(meow_id))
meow_to_rgn <- read_csv(file.path(dir_git, '_spatial/iucn_rgn_to_meow.csv'))

```

### Loop over turtle species

``` {r loop comprehensively assessed}
### Make a list of species to loop over:

turtle_spp <- c('Caretta', 'Dermochelys', 'Lepidochelys', 'Eretmochelys', 'Chelonia', 'Natator') %>%
  paste0(collapse = '|')
turtle_maps <- read_csv(file.path(dir_data, 'spp_marine_maps_2018-1.csv')) %>%
  filter(str_detect(sciname, turtle_spp))

spp_names <- turtle_maps$sciname %>% 
  unique()

reload <- FALSE

##########################################################.
### Looping over taxonomic groups -----
##########################################################.
for(spp_name in spp_names) {
  ### spp_name <- spp_names[1]
  spp_name_clean <- spp_name %>%
    tolower() %>%
    str_replace_all('[^a-z]', '_')

  spp_sum_file <- file.path(dir_git, 'pres_figs', sprintf('rast/%s_risk.csv', spp_name_clean))
  
  if(!file.exists(spp_sum_file) | reload) {
    ### Filter to just this taxon, and only assessed within this taxon
    spp_maps <- turtle_maps %>%
      filter(str_detect(sciname, spp_name)) %>%
      filter(comp_assessed == TRUE)
    
    spp_risk_trend_sum <- spp_risk_trend %>%
      filter(iucn_sid %in% spp_maps$iucn_sid) %>%
      filter(!is.na(cat_score)) %>%
      arrange(iucn_sid)
    
    ### Using the iucn_sid field, generate a vector of all species range files for
    ### this taxon.
    spp_ids <- spp_risk_trend_sum$iucn_sid %>%
      unique()
    
    cat_msg('Processing ', length(spp_ids),
            ' comp-assessed species maps in ', spp_name, '...')

    ##########################################################.
    ### Looping over species within group -----
    ##########################################################.
    ### Collect all species ranges for this spp into a single data.frame.
    ### Use mclapply since we're reading many large-ish files.  For MARINE_MAMMALS (85 assessed spp)
    ### this takes about 30-40 seconds
    spp_cells_list <- parallel::mclapply(spp_ids, mc.cores = 24,
                                        FUN = function(x) {
        ### x <- spp_ids[1]
        ### x <- 6336                            
        csv_file <- file.path(dir_o_anx, 'spp_rasters', 
                              sprintf('iucn_sid_%s.csv', x))
        
        spp_risk_map <- read_csv(csv_file, col_types = 'di') %>%
          mutate(iucn_sid = x) %>%
          left_join(spp_risk_trend_sum, by = 'iucn_sid') 
        
        ### Identify regional assessments if any
        meow_rgns <- meow_to_rgn %>%
          filter(iucn_rgn %in% spp_risk_map$iucn_rgn)
        
        non_global_rgns <- meow_rgns %>%
          filter(iucn_rgn != 'global')
  
        if(nrow(non_global_rgns) > 0) {
          ### If any regional assessments, clip the MEOW cells down to the appropriate region...
          meow_cells <- meow_cells_all %>%
            inner_join(meow_rgns, by = 'meow_id') %>%
            rename(rgn_name = iucn_rgn)
          ### ... then filter out non-matching overlapped cells
          spp_risk_map <- spp_risk_map %>%
            left_join(meow_cells, by = 'cell_id') %>%
            mutate(rgn_name = ifelse(is.na(meow_id), 'global', rgn_name),
                   priority = ifelse(rgn_name == 'global', 100, priority)) %>%
            filter(iucn_rgn == rgn_name) %>%
            group_by(cell_id) %>%
            filter(priority == min(priority)) %>%
            ungroup()
          ### NOTE: at this point, still possible to have multiple regional
          ### assessments, if priorities are the same (e.g. Europe and Pan Africa,
          ### or overlaps around Africa).  Those values will be averaged in
          ### the group_by() below.
        }
        
        ### select down to main columns; also, if presence == 5 (extinct), 
        ### adjust category and trend scores.
        spp_risk_sum <- spp_risk_map %>%
          select(cell_id, presence, iucn_sid, iucn_rgn, cat_score, trend_score) %>%
          distinct() %>%
          mutate(cat_score   = ifelse(presence == 5, 1, cat_score),        
                   ### fix local extinctions
                 trend_score = ifelse(presence == 5, NA, trend_score)) # %>% 
                   ### fix local extinctions
          # group_by(cell_id, iucn_sid) %>%
          # summarize(cat_score   = mean(cat_score, na.rm = TRUE),
          #           trend_score = mean(trend_score, na.rm = TRUE)) %>%
          # ungroup() %>%
          # mutate(trend_score = ifelse(is.nan(trend_score), NA, trend_score))
        
        return(spp_risk_sum)
      }) ### end of mclapply over all species in taxonomic group
  
    ##########################################################.
    ### Processing cell calculations for group -----
    ##########################################################.
  
    ### Set up for keyed data.table merging: key for iucn_sid, cell_id
    cat_msg('...binding rows to data.frame...')

    spp_risk_map <- bind_rows(spp_cells_list) ### this is pretty fast!
    
    cat_msg('...summarizing...')
    spp_risk_summary <- spp_risk_map %>%
      group_by(cell_id) %>%
      summarize(mean_risk   = mean(cat_score),
                  ### NA categories already filtered out
                # var_risk    = var(cat_score),
                # n_spp_threatened = sum(cat_score >= 0.4 & cat_score < 1),  
                n_spp_risk  = n()) %>%
      ungroup()
                # mean_trend  = mean(trend_score, na.rm = TRUE), 
                  ### not every species has a trend; na.rm = TRUE
                # mean_trend  = ifelse(is.nan(mean_trend), NA, mean_trend),
                # n_spp_trend = sum(!is.na(trend_score)))
    
    cat_msg('...writing file', spp_sum_file, '...')
    write_csv(spp_risk_summary, spp_sum_file)

  } else { ### end of if statement checking whether file exists for this taxon
    cat_msg('Found file ', spp_sum_file, '... skipping process...')
  }
  
} ### end of taxonomic group loop
  
```
