---
title: 'MPAs and biodiversity intactness'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohibc/src/templates/ohibc_hdr1.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(raster)
library(data.table)

source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')
  ### includes library(tidyverse); library(stringr); 
  ### dir_M points to ohi directory on Mazu; dir_O points to home dir on Mazu

dir_git <- '~/github/spp_risk_dists'

### goal specific folders and info
dir_data  <- file.path(dir_git, '_data')
dir_o_anx <- file.path(dir_O, 'git-annex/spp_risk_dists')

source(file.path(dir_git, '_setup/common_fxns.R'))

```

## Compare MPA coverage to biodiversity risk

MPA coverage will be compared to non-MPA areas within EEZs.  The .csv codes MPAs as their numeric IUCN category (i.e. Ia = Ib = 1, II = 2, ..., VI = 6); areas that are specifically designated as no-take (that are not already in categories Ia, Ib, and II) are coded as -1; areas that are designated in the WDPA database but do not fall into one of these categories are coded as 8.

``` {r set up mpa and values dataframe}

mean_rast     <- raster(file.path(dir_git, '_output', 'mean_risk_raster_comp.tif'))
eez_rast      <- raster(file.path(dir_git, '_spatial', 'eez_rast.tif'))
ocean_area_rast <- raster(file.path(dir_git, '_spatial', 'ocean_area_rast.tif'))

risk_df <- data.frame(cell_id   = 1:length(values(mean_rast)),
                      ocean_area = values(ocean_area_rast),
                      eez       = values(eez_rast),
                      mean_risk = values(mean_rast)) %>%
  filter(!is.na(mean_risk))

```

``` {r set up no take vs risk df}

calc_mpa_areas <- function(df) {
  ### ensure MPA area is no larger than ocean area of the cell; and
  ### code NA as zero for means and weights calculations
  df %>%
    mutate(mpa = ifelse(is.na(mpa), 0, mpa),
           mpa = ifelse(mpa > ocean_area, ocean_area, mpa),
           non_mpa = ocean_area - mpa) %>%
    gather(mpa_status, area, mpa, non_mpa) %>%
    mutate(mpa_status = (mpa_status == 'mpa'))
}

mpa_df   <- read_csv(file.path(dir_git, '_spatial', 'wdpa_mpa_area.csv'),
                          col_types = 'dddd') %>%
  filter(wdpa_category <= 2) %>%
  select(-wdpa_category) %>%
  group_by(cell_id) %>%
  summarize(mpa = sum(prot_area_km2, na.rm = TRUE))

risk_mpa_df <- risk_df %>%
  left_join(mpa_df, by = 'cell_id') %>%
  calc_mpa_areas()

```


``` {r no take mpas, eval = TRUE}

global_no_take <- risk_mpa_df %>%
  filter(eez < 255 & eez != 213) %>%
  mutate(lbl = 'Global EEZ')


highseas_no_take <- risk_mpa_df %>%
  filter(eez > 255 | eez == 213) %>%
  mutate(lbl = 'High seas')

no_take_df <- bind_rows(global_no_take, highseas_no_take) %>%
  mutate(lbl = forcats::fct_inorder(lbl))

prot_pct <- no_take_df %>%
  group_by(lbl) %>%
  summarize(pct_prot = sum(area * mpa_status) / sum(area) * 100,
            tot_area = sum(area))
### this is independent of range-rarity weighting; since `area` includes only
### ocean area this calculation accounts for percent of _marine_ area protected
### by MPAs.

no_take_means <- no_take_df %>%
  group_by(mpa_status, lbl) %>%
  summarize(mu = sum(mean_risk * area) / sum(area)) %>%
  ungroup()
  
```


## Plotting Mean risk

To capture area of partially protected cells, use `geom_density(aes(weight = area, x = mean_risk, ..scaled..))`

``` {r plot_function}
plot_dist <- function(df, means_df, var_name, x_lims = c(0, .6)) {
  x <- ggplot(df, aes(x = mean_risk)) +
    ggtheme_plot(base_size = 7) +
    theme(strip.text.y = element_text(angle = 0, hjust = 0),
          axis.text.y  = element_blank(),
          axis.title   = element_blank(),
          panel.grid.major.y = element_blank(),
          plot.margin = unit(c(.1, .25, .1, .35), 'cm'),
          legend.position = c(.75, .55)) +
    geom_density(aes(x = mean_risk, weight = area, ..scaled..,
                     fill = mpa_status, color = mpa_status), 
                 alpha = .5, size = .25) +
    geom_vline(data = means_df, 
               aes(xintercept = mu, linetype = mpa_status),
               color = 'grey20', alpha = .8) +
    scale_x_continuous(expand = c(0, 0), 
                       limits = x_lims,
                       labels = risk_lbls,
                       breaks = risk_brks) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_fill_manual(values  = c('red4', 'cadetblue2')) +
    scale_color_manual(values = c('red3', 'cadetblue4')) +
    labs(fill = var_name, color = var_name, linetype = var_name)
  
  return(x)
}

```

``` {r generate figure}

prot_pct_lbls <- prot_pct %>%
  mutate(area_lbl = sprintf('%.1f x 10⁶ km²', tot_area/1e6),
         prot_lbl = sprintf('%.1f%% no take', pct_prot))

mpa_plot_eez_prot_only <- plot_dist(no_take_df %>% 
                            filter(lbl == 'Global EEZ') %>%
                            filter(mpa_status), 
                          no_take_means %>% 
                            filter(lbl == 'Global EEZ') %>%
                            filter(mpa_status),  
                          'Marine reserve') +
    scale_fill_manual(values  = c('cadetblue2')) +
    scale_color_manual(values = c('cadetblue4')) +
    scale_linetype_manual(values = c('dashed'))

ggsave(file.path(dir_git, 'pres_figs/fig/mpa_plot_global_eez_prot_only.png'),
       height = 1, width = 7.25, dpi = 300, units = 'in')

mpa_plot_eez <- plot_dist(no_take_df %>% filter(lbl == 'Global EEZ'), 
                          no_take_means %>% filter(lbl == 'Global EEZ'), 
                          'Marine reserve')
ggsave(file.path(dir_git, 'pres_figs/fig/mpa_plot_global_eez.png'),
       height = 1, width = 7.25, dpi = 300, units = 'in')

mpa_plot_hs  <- plot_dist(no_take_df %>% filter(lbl == 'High seas'), 
                          no_take_means %>% filter(lbl == 'High seas'),
                          'Marine reserve')
ggsave(file.path(dir_git, 'pres_figs/fig/mpa_plot_high_seas.png'),
       height = 1, width = 7.25, dpi = 300, units = 'in')
```


