---
title: 'Explore hotspots'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/ohibc/src/templates/ohibc_hdr1.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(raster)
library(data.table)

source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')
  ### includes library(tidyverse); library(stringr); 
  ### dir_M points to ohi directory on Mazu; dir_O points to home dir on Mazu

dir_git <- '~/github/spp_risk_dists'

### goal specific folders and info
dir_setup <- file.path(dir_git, 'data_setup')
dir_data  <- file.path(dir_git, 'data')
dir_o_anx <- file.path(dir_O, 'git-annex/spp_risk_dists')

source(file.path(dir_git, 'data_setup/api_fxns.R'))

```

# Summary

Having selected some cells from regions of interest, explore these by mapping and comparing values.  Identify species found within each cell.

## Identify species groups

Using the summary files, identify the species *groups* present in each cell of interest.

Then get the processed rasters to get values for n_spp, risk, variance, and rr-weighted risk/var.  Also get eez values and depths.


``` {r compare_sum_files_to_interest_cells}

cells_of_interest <- readxl::read_excel(file.path(dir_git, 'data_explore', 'cells_to_explore.xlsx'))

cell_summary_file <- file.path(dir_git, 'data_explore', 'cell_sums_of_interest.csv')

reload <- FALSE

if(!file.exists(cell_summary_file) | reload == TRUE) {
  dir_taxa_summaries <- file.path(dir_o_anx, 'taxa_summaries')
  sum_files <- list.files(dir_taxa_summaries,
                          pattern = sprintf('cell_sum_%s.csv', api_version),
                          full.names = TRUE)
  message('reading summary files')
  ptm <- system.time({
    cell_values_all <- parallel::mclapply(sum_files, mc.cores = 32,
                                          FUN = function(x) { ### x <- sum_files[1]
                                            read_csv(x, col_types = 'dddiidi') %>%
                                              mutate(spp_gp = basename(x) %>% 
                                                       str_replace('_cell_sum.+', ''))
                                            }) %>%
      bind_rows()
  }) ### end of system.time
  message('... processing time ', ptm[3], ' sec')
  
  cells_of_interest_sum <- cell_values_all %>%
    filter(cell_id %in% cells_of_interest$cell_id) %>%
    select(cell_id, n_spp_gp = n_spp_risk, spp_gp) %>%
    left_join(cells_of_interest, by = 'cell_id') %>%
    arrange(cell_group, cell_id)
  
  cell_df <- data.frame(cell_id = values(raster(file.path(dir_git, 'output', 
                                                          'cell_id_raster.tif'))),
                        n_spp   = values(raster(file.path(dir_git, 'output',
                                                          'n_spp_risk_raster_010deg.tif'))),
                        risk    = values(raster(file.path(dir_git, 'output',
                                                          'mean_risk_raster_010deg.tif'))),
                        rr_risk = values(raster(file.path(dir_git, 'output', 
                                                          'mean_rr_risk_raster_010deg.tif'))),
                        rr_var  = values(raster(file.path(dir_git, 'output', 
                                                          'var_rr_risk_raster_010deg.tif'))),
                        eez_id  = values(raster(file.path(dir_git, 'spatial', 
                                                          'eez_rast_010_wgs84.tif'))),
                        lme_id  = values(raster(file.path(dir_git, 'spatial', 
                                                          'lme_rast_010_wgs84.tif'))),
                        depth   = values(raster(file.path(dir_git, 'spatial', 
                                                          'bathy_rast_010_wgs84.tif')))) %>%
    filter(cell_id %in% cells_of_interest$cell_id)
  
  cells_info <- cells_of_interest_sum %>%
    full_join(cell_df, by = 'cell_id')
  
  write_csv(cells_info, cell_summary_file)
}
  
```

## Get all the species in each cell using the original raster extracts

``` {r get_spp_in_cells_of_interest}

cells_info <- read_csv(cell_summary_file)

# spp_gps_all <- cells_info$spp_gp %>% unique()
  ### all are represented here...

spp_maps <- read_csv(file.path(dir_data, sprintf('spp_marine_maps_%s.csv', api_version)),
                     col_types = 'ddciccc')

taxa_cells_file <- file.path(dir_git, 'data_explore', 'taxa_spp_cells.csv')

reload <- FALSE

if(!file.exists(taxa_cells_file) | reload == TRUE) {
  ### Make a list of taxonomic groups to loop over:
  taxa <- spp_maps$dbf_file %>%
    unique() %>%
    str_replace('\\....$', '')
  
  taxa_cells_list <- vector('list', length = length(taxa))
  
  for(i in seq_along(taxa)) { ### i <- 5
    taxon <- taxa[i]
    
    spp_ids_in_taxon <- spp_maps %>%
      filter(str_detect(dbf_file, taxon)) %>%
      .$iucn_sid
    cat(sprintf('processing %s spp in %s...\n', length(spp_ids_in_taxon), taxon))
    
    spp_cells <- parallel::mclapply(spp_ids_in_taxon, mc.cores = 32,
      FUN = function(x) { ### x <- spp_ids_in_taxon[1]
        f <- file.path(dir_o_anx, 'spp_rasters',
                       sprintf('iucn_sid_%s_010deg.csv', x))
        if(file.exists(f)) {
          y <- read_csv(f, col_types = 'di') %>%
            mutate(iucn_sid = x) %>%
            select(-presence)  %>%
            filter(cell_id %in% cells_info$cell_id)
        } else {
          y <- data.frame(cell_id = NA,
                          iucn_sid = x, 
                          f = f, error = 'file not found')
        }
        return(y)
      }) %>%
      bind_rows() %>%
      mutate(spp_gp = taxon)
    
    taxa_cells_list[[i]] <- spp_cells
  }
  
  taxa_cells_df <- taxa_cells_list %>%
    bind_rows()  %>%
    filter(!is.na(cell_id)) %>%
    select(cell_id, iucn_sid, spp_gp) %>%
    left_join(spp_maps %>% select(iucn_sid, max_depth), by = 'iucn_sid') %>%
    distinct() 
  
  write_csv(taxa_cells_df, taxa_cells_file)
}

```

``` {r attach_spp_cell_info}

spp_risk <- read_csv(file.path(dir_data, sprintf('iucn_risk_current_%s.csv', api_version)),
                     col_types = 'dc_____d_') %>%
  distinct()

spp_risk_rgn <- read_csv(file.path(dir_data, sprintf('iucn_risk_rgn_current_%s.csv', api_version)),
                         col_types = 'dc_cd_') %>%
  distinct()

spp_ranges   <- read_csv(file.path(dir_data, sprintf('iucn_spp_range_area_%s.csv', api_version)),
                         col_types = 'dd__') %>%
  distinct()

### make a dataframe of species risk and regional risk
spp_risk_all <- spp_risk %>%
  mutate(iucn_rgn = 'global') %>%
  bind_rows(spp_risk_rgn) %>%
  left_join(spp_ranges, by = c('iucn_sid')) %>%
  select(iucn_sid, sciname, iucn_rgn, cat_score, range_km2) %>%
  mutate(range_km2 = round(range_km2))

cells_info <- read_csv(cell_summary_file, col_types = 'd__cd____id') %>%
  distinct()

lme_to_rgn <- read_csv(file.path(dir_git, 'spatial/iucn_rgn_to_lme.csv')) %>%
  rename(rgn_name = iucn_rgn) %>%
  distinct()

taxa_cells_info <- read_csv(taxa_cells_file, col_types = 'ddcc') %>%
  left_join(spp_risk_all, by = 'iucn_sid') %>%
  mutate(spp_gp = tolower(spp_gp)) %>%
  left_join(cells_info, by = c('cell_id')) %>%
    ### add info from rasters, including lme_id
  left_join(lme_to_rgn, by = c('lme_id')) %>%
    ### add region names to filter out regional assessments
  mutate(rgn_name = ifelse(is.na(lme_id), 'global', rgn_name),
         priority = ifelse(rgn_name == 'global', 100, priority)) %>%
    ### LME cells already tagged 'global'; fix non-LME cells  and set low-ranking priority
  filter(iucn_rgn == rgn_name) %>%
  group_by(cell_id, iucn_sid) %>%
  filter(priority == min(priority)) %>%
    ### for each spp in each cell, choose the obs with highest-ranking priority
  ungroup()

spp_cells_info <- taxa_cells_info %>%
  select(cell_id, iucn_sid, sciname, spp_gp, iucn_rgn, 
         cat_score, range_km2, cell_group, depth, spp_max_depth = max_depth)

write_csv(spp_cells_info, file.path(dir_git, 'data_explore', 'spp_cells_scores.csv'))

```

## Display points on a map and show the species list

For each cell group, plot a map of range-rarity-weighted risk; display the cells for that region on the map; 
``` {r create_global_map_w_points}

spp_cells_info <- read_csv(file.path(dir_git, 'data_explore', 'spp_cells_scores.csv'),
                           col_types = 'ddcccddcdc') %>%
  arrange(cell_group)

land_poly <- sf::read_sf(file.path(dir_git, 'spatial/ne_10m_land/ne_10m_land.shp'))


cell_id_file <- file.path(dir_git, 'data_explore/cell_id_latlong.csv')
if(!file.exists(cell_id_file)) {
  cell_id_df     <- raster(file.path(dir_git, 'output', 'cell_id_raster.tif')) %>%
    rasterToPoints() %>%
    as.data.frame() %>%
    setNames(c('x', 'y', 'cell_id')) %>%
    filter(cell_id %in% spp_cells_info$cell_id) %>%
    write_csv(cell_id_file)
} else {
  cell_id_df <- read_csv(cell_id_file)
}

fig_file <- file.path(dir_git, 'data_explore/figs',
                    sprintf('global_map_pts_of_interest.png'))

message(sprintf('Mapping cells for global cells of interest'))

### Attach lat/long values to cell IDs
groups_latlong <- spp_cells_info %>% 
  select(cell_id, cell_group) %>% 
  distinct() %>%
  left_join(cell_id_df, by = 'cell_id') %>%
  group_by(cell_group) %>%
  summarize(x = mean(x), y = mean(y))
  
if(!file.exists(fig_file)) {
  ### create the image and save it
  
  if(!exists('rr_risk_df')) {
    rr_risk_df <- raster(file.path(dir_git, 'output', 'mean_rr_risk_raster_010deg.tif')) %>%
      rasterToPoints() %>%
      as.data.frame() %>%
      setNames(c('x', 'y', 'rr_risk'))
  }
  
  x <- ggplot(rr_risk_df) +
    ggtheme_map() +
    geom_raster(aes(x, y, fill = rr_risk)) +
    geom_sf(data = land_poly, aes(geometry = geometry), fill = 'grey96', color = 'grey40', size = .10) +
    geom_point(data = groups_latlong, aes(x, y), 
               shape = 21, color = 'white', fill = 'white', size = 2.5) +
    geom_point(data = groups_latlong, aes(x, y), 
               shape = 21, color = 'blue', fill = 'yellow', size = 2) +
    scale_fill_gradientn(colors = c('green4', 'lightyellow', 'red2', 'red3', 'red4', 'purple4'),
                         limits = c(0, 1),
                         labels = c('LC', 'NT', 'VU', 'EN', 'CR', 'EX'),
                         breaks = c( 0.0,  0.2,  0.4,  0.6,  0.8,  1.0)) +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(expand = c(0, 0)) +
    labs(title = 'Global areas of interest',
         fill  = 'risk')
  
  ggsave(plot = x, file = fig_file,
         width = 6, height = 4, units = 'in', dpi = 300)
}

knitr::include_graphics(path.expand(fig_file))

```

### Region maps with spp data

``` {r create_rgn_maps_w_points}

cell_gps <- unique(spp_cells_info$cell_group)

spp_cell_gps <- vector('list', length = length(cell_gps)) %>%
  setNames(cell_gps)
fig_files <- vector('list', length = length(cell_gps)) %>%
  setNames(cell_gps)

for(cell_gp in cell_gps) {
  ### cell_gp <- cell_gps[1]
  
  fig_file <- file.path(dir_git, 'data_explore/figs',
                      sprintf('rgn_map_%s.png', 
                              tolower(cell_gp) %>% str_replace_all('[^a-z0-9]', '_')))

  message(sprintf('Mapping cells for area of interest: %s\n', cell_gp))

  ### Attach lat/long values to cell IDs
  cells_latlong <- spp_cells_info %>% 
    filter(cell_group == cell_gp) %>%
    select(cell_id, depth) %>% 
    distinct() %>%
    left_join(cell_id_df, by = 'cell_id') %>%
    arrange(cell_id) %>%
    mutate(cell_name = toupper(letters[1:n()]))
  
  spp_cell_gps[[cell_gp]] <- spp_cells_info %>%
    filter(cell_group == cell_gp) %>%
    left_join(cells_latlong, by = c('cell_id', 'depth')) %>%
    select(-x, -y, -cell_group, -cell_id, -depth) %>%
    mutate(present = TRUE) %>%
    spread(cell_name, present, fill = FALSE)

  if(!file.exists(fig_file)) {
    ### create the image and save it
    
    if(!exists('rr_risk_df')) {
      rr_risk_df <- raster(file.path(dir_git, 'output', 'mean_rr_risk_raster_010deg.tif')) %>%
        rasterToPoints() %>%
        as.data.frame() %>%
        setNames(c('x', 'y', 'rr_risk'))
    }
    
    ### define bounding box with some margin around it:
    bbox <- c(xmin = max(min(cells_latlong$x) - 4, -179.95),
              xmax = min(max(cells_latlong$x) + 4,  179.95),
              ymin = max(min(cells_latlong$y) - 2,  -89.95),
              ymax = min(max(cells_latlong$y) + 2,   89.95))
    
    depth_label <- sprintf('Depths: %s', 
                           paste0(cells_latlong$cell_name, ': ', 
                                  cells_latlong$depth, ' m', 
                                  collapse = '; '))
      
    x <- ggplot(rr_risk_df) +
      ggtheme_map() +
      geom_raster(aes(x, y, fill = rr_risk)) +
      geom_sf(data = land_poly, aes(geometry = geometry), fill = 'grey96', color = 'grey40', size = .10) +
      geom_point(data = cells_latlong, aes(x, y), 
                 shape = 21, color = 'blue', fill = 'yellow', size = 2) +
      geom_text(data = cells_latlong, aes(x, y, label = cell_name),
                nudge_x = .15, nudge_y = .15) +
      scale_fill_gradientn(colors = c('green4', 'lightyellow', 'red2', 'red3', 'red4', 'purple4'),
                           limits = c(0, 1),
                           labels = c('LC', 'NT', 'VU', 'EN', 'CR', 'EX'),
                           breaks = c( 0.0,  0.2,  0.4,  0.6,  0.8,  1.0)) +
      annotate('text', x = bbox[1] + .05, y = bbox[3] + .05,
               label = depth_label, hjust = 0, vjust = 0) +
      coord_sf(xlim = c(bbox[1], bbox[2]),  ylim = c(bbox[3], bbox[4])) +
      scale_x_continuous(expand = c(0, 0)) +
      scale_y_continuous(expand = c(0, 0)) +
      labs(title = cell_gp,
           fill  = 'mean risk')
    
    ggsave(plot = x, file = fig_file,
           width = 6, height = 4, units = 'in', dpi = 300)
  }
  
  fig_files[[cell_gp]] <- fig_file

}
```

-----

``` {r}

cell_gp <- cell_gps[1]

knitr::include_graphics(path.expand(fig_files[[cell_gp]]))

DT::datatable(spp_cell_gps[[cell_gp]], caption = cell_gp)
```

-----

``` {r}

cell_gp <- cell_gps[2]

knitr::include_graphics(path.expand(fig_files[[cell_gp]]))

DT::datatable(spp_cell_gps[[cell_gp]], caption = cell_gp)
```

-----

``` {r}

cell_gp <- cell_gps[3]

knitr::include_graphics(path.expand(fig_files[[cell_gp]]))

DT::datatable(spp_cell_gps[[cell_gp]], caption = cell_gp)
```

-----

``` {r}

cell_gp <- cell_gps[4]

knitr::include_graphics(path.expand(fig_files[[cell_gp]]))

DT::datatable(spp_cell_gps[[cell_gp]], caption = cell_gp)
```

-----

``` {r}

cell_gp <- cell_gps[5]

knitr::include_graphics(path.expand(fig_files[[cell_gp]]))

DT::datatable(spp_cell_gps[[cell_gp]], caption = cell_gp)
```

-----

``` {r}

cell_gp <- cell_gps[6]

knitr::include_graphics(path.expand(fig_files[[cell_gp]]))

DT::datatable(spp_cell_gps[[cell_gp]], caption = cell_gp)
```

-----

``` {r}

cell_gp <- cell_gps[7]

knitr::include_graphics(path.expand(fig_files[[cell_gp]]))

DT::datatable(spp_cell_gps[[cell_gp]], caption = cell_gp)
```

-----

``` {r}

cell_gp <- cell_gps[8]

knitr::include_graphics(path.expand(fig_files[[cell_gp]]))

DT::datatable(spp_cell_gps[[cell_gp]], caption = cell_gp)
```

-----

``` {r}

cell_gp <- cell_gps[9]

knitr::include_graphics(path.expand(fig_files[[cell_gp]]))

DT::datatable(spp_cell_gps[[cell_gp]], caption = cell_gp)
```

-----

``` {r}

cell_gp <- cell_gps[10]

knitr::include_graphics(path.expand(fig_files[[cell_gp]]))

DT::datatable(spp_cell_gps[[cell_gp]], caption = cell_gp)
```

-----

``` {r}

cell_gp <- cell_gps[11]

knitr::include_graphics(path.expand(fig_files[[cell_gp]]))

DT::datatable(spp_cell_gps[[cell_gp]], caption = cell_gp)
```

-----

``` {r}

cell_gp <- cell_gps[12]

knitr::include_graphics(path.expand(fig_files[[cell_gp]]))

DT::datatable(spp_cell_gps[[cell_gp]], caption = cell_gp)
```

-----

``` {r}

cell_gp <- cell_gps[13]

knitr::include_graphics(path.expand(fig_files[[cell_gp]]))

DT::datatable(spp_cell_gps[[cell_gp]], caption = cell_gp)
```

-----

``` {r}

cell_gp <- cell_gps[14]

knitr::include_graphics(path.expand(fig_files[[cell_gp]]))

DT::datatable(spp_cell_gps[[cell_gp]], caption = cell_gp)
```

-----

``` {r}

cell_gp <- cell_gps[15]

knitr::include_graphics(path.expand(fig_files[[cell_gp]]))

DT::datatable(spp_cell_gps[[cell_gp]], caption = cell_gp)
```

-----

``` {r}

cell_gp <- cell_gps[16]

knitr::include_graphics(path.expand(fig_files[[cell_gp]]))

DT::datatable(spp_cell_gps[[cell_gp]], caption = cell_gp)
```

-----

``` {r}

cell_gp <- cell_gps[17]

knitr::include_graphics(path.expand(fig_files[[cell_gp]]))

DT::datatable(spp_cell_gps[[cell_gp]], caption = cell_gp)
```

-----

``` {r}

cell_gp <- cell_gps[18]

knitr::include_graphics(path.expand(fig_files[[cell_gp]]))

DT::datatable(spp_cell_gps[[cell_gp]], caption = cell_gp)
```


