---
title: 'Regress risk and trend vs. geographic info'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/src/templates/ohara_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(raster)

source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')

dir_git <- '~/github/spp_health_dists'

### goal specific folders and info
dir_data    <- file.path(dir_git, 'data')
dir_spatial <- file.path(dir_git, 'spatial')
dir_anx     <- file.path(dir_M, 'git-annex')
dir_o_anx   <- file.path(dir_O, 'git-annex/spp_health_dists')
# ### provenance tracking
# library(provRmd); prov_setup()

### support scripts
# source('https://raw.githubusercontent.com/oharac/src/master/R/rast_tools.R')) 
  ### raster plotting and analyzing scripts

```

# Summary

At the LOICZID raster scale (0.5° cells), compare ecosystem risk and trends to geographic features.

# Data Sources

# Methods {.tabset}

## Risk/trend by cell

``` {r check_dists}

risk_df <- read_csv(file.path(dir_data, 'current_risk_by_loiczid.csv'),
                    col_types = 'ddddd') %>% 
  filter(n_spp > 0) %>%
  mutate(log_n_spp = log(n_spp))

x <- risk_df %>% filter(mean_risk == 0)

ggplot(x, aes(x = n_spp)) +
  ggtheme_plot() +
  geom_histogram() +
  labs(title = 'n_spp for (mean_risk == 0)')

ggplot(risk_df, aes(x = mean_risk)) +
  ggtheme_plot() +
  geom_histogram() +
  labs(title = 'mean_risk across all cells')

ggplot(risk_df, aes(x = log_mean_risk)) +
  ggtheme_plot() +
  geom_histogram() +
  labs(title = 'log(mean_risk) across all non-NA cells')

# ggplot(risk_df %>% filter(n_spp > 0), aes(x = n_spp)) +
#   ggtheme_plot() +
#   geom_histogram() +
#   labs(title = 'number of species by cell')
# 
# ggplot(risk_df, aes(x = log_n_spp)) +
#   ggtheme_plot() +
#   geom_histogram()

trend_df <- read_csv(file.path(dir_data, 'spp_trend_by_loiczid.csv'),
                    col_types = 'ddd')

ggplot(trend_df, aes(x = mean_trend)) +
  ggtheme_plot() +
  geom_histogram()

```

Note: For cells with at least 5 species identified, un-logged-risk metric provides `r nrow(risk_df %>% filter(n_spp >= 5))` observations (cells), `r nrow(risk_df %>% filter(n_spp >= 5 & mean_risk == 0))` of which are zeros.  The logged-risk metric provides `r nrow(risk_df %>% filter(n_spp >= 5 & !is.na(log_mean_risk)))` observations (after dropping zeros).

### Mean risk:

``` {r}
x <- qqnorm(risk_df$mean_risk, main = 'Normal Q-Q, mean risk', ylab = 'mean risk')
```

### Log (mean risk):

``` {r}
x <- qqnorm(risk_df$mean_risk, main = 'Normal Q-Q, log(mean risk)', ylab = 'log (mean risk)')
```

### Mean trend:

``` {r}
x <- qqnorm(risk_df$mean_risk, main = 'Normal Q-Q, mean trend', ylab = 'mean trend')
```


## Geographic vars

``` {r scatter_plots}

risk_df <- read_csv(file.path(dir_data, 'current_risk_by_loiczid.csv'),
                    col_types = 'ddddd')

trend_df <- read_csv(file.path(dir_data, 'spp_trend_by_loiczid.csv'),
                    col_types = 'ddd') %>%
  rename(n_trend_spp = n_spp)

lat_df <- read_csv(file.path(dir_data, 'latitude_lookup.csv'), col_types = 'dd')

ocean_area_df <- read_csv(file.path(dir_data, 'ocean_area_lookup.csv'), col_types = 'dd')

area_by_lat <- lat_df %>% 
  full_join(ocean_area_df) %>%
  group_by(lat) %>%
  summarize(lat_area = sum(ocean_area_km2, na.rm = TRUE)) %>%
  ungroup()

risk_trend_df <- risk_df %>%
  full_join(trend_df, by = 'loiczid') %>%
  filter(n_spp >= 5) %>%
  full_join(lat_df,   by = 'loiczid') %>%
  mutate(lat_abs = abs(lat))

ggplot(risk_trend_df, aes(x = mean_risk, y = mean_trend)) +
  ggtheme_plot() +
  geom_point(alpha = .01) +
  stat_smooth(method = 'lm') +
  labs(title = 'Mean risk vs mean trend')

ggplot(risk_trend_df, aes(x = log_mean_risk, y = mean_trend)) +
  ggtheme_plot() +
  geom_point(alpha = .01) +
  stat_smooth(method = 'lm') +
  labs(title = 'Log(mean risk) vs mean trend')

ggplot(risk_trend_df, aes(x = lat, y = mean_risk)) +
  ggtheme_plot() +
  geom_point(alpha = .01) +
  stat_smooth(method = 'lm') +
  labs(title = 'Mean risk vs latitude')

ggplot(risk_trend_df, aes(x = lat, y = log_mean_risk)) +
  ggtheme_plot() +
  geom_point(alpha = .01) +
  stat_smooth(method = 'lm') +
  labs(title = 'Log(mean risk) vs latitude')

ggplot(risk_trend_df, aes(x = lat, y = mean_trend)) +
  ggtheme_plot() +
  geom_point(alpha = .01) +
  stat_smooth(method = 'lm') +
  labs(title = 'Mean trend vs latitude')

ggplot(risk_trend_df, aes(x = lat_abs, y = mean_risk)) +
  ggtheme_plot() +
  geom_point(alpha = .01) +
  stat_smooth(method = 'lm') +
  labs(title = 'Log(mean risk) vs abs(latitude)')

ggplot(risk_trend_df, aes(x = lat_abs, y = log_mean_risk)) +
  ggtheme_plot() +
  geom_point(alpha = .01) +
  stat_smooth(method = 'lm') +
  labs(title = 'Log(mean risk) vs abs(latitude)')

ggplot(risk_trend_df, aes(x = lat_abs, y = mean_trend)) +
  ggtheme_plot() +
  geom_point(alpha = .01) +
  stat_smooth(method = 'lm') +
  labs(title = 'Log(mean risk) vs abs(latitude)')

ggplot(area_by_lat, aes(x = lat, y = lat_area)) +
  ggtheme_plot() +
  geom_area(fill = 'blue4', alpha = .3)


```

## Regressions

``` {r regressions}

risk_df <- read_csv(file.path(dir_data, 'current_risk_by_loiczid.csv'),
                    col_types = 'ddddd')

trend_df <- read_csv(file.path(dir_data, 'spp_trend_by_loiczid.csv'),
                    col_types = 'ddd') %>%
  rename(n_trend_spp = n_spp)

lat_df <- read_csv(file.path(dir_data, 'latitude_lookup.csv'), col_types = 'dd')

ocean_area_df <- read_csv(file.path(dir_data, 'ocean_area_lookup.csv'), col_types = 'dd')

risk_trend_df <- risk_df %>%
  full_join(trend_df, by = 'loiczid') %>%
  filter(n_spp >= 5) %>%
  full_join(lat_df,   by = 'loiczid') %>%
  mutate(lat_abs = abs(lat))

risk_trend_mdl1 <- lm(mean_risk ~ mean_trend,
                data = risk_trend_df)
summary(risk_trend_mdl1)

risk_trend_mdl2 <- lm(log_mean_risk ~ mean_trend,
                data = risk_trend_df)
summary(risk_trend_mdl2)

risk_geog_mdl1 <- lm(mean_risk ~ lat,
                     data = risk_trend_df)
summary(risk_geog_mdl1)

risk_geog_mdl2 <- lm(log_mean_risk ~ lat,
                     data = risk_trend_df)
summary(risk_geog_mdl2)

risk_geog_mdl3 <- lm(mean_risk ~ lat_abs,
                     data = risk_trend_df)
summary(risk_geog_mdl3)

risk_geog_mdl4 <- lm(log_mean_risk ~ lat_abs,
                     data = risk_trend_df)
summary(risk_geog_mdl4)

trend_geog_mdl1 <- lm(mean_trend ~ lat,
                     data = risk_trend_df)
summary(trend_geog_mdl1)

trend_geog_mdl2 <- lm(mean_trend ~ lat_abs,
                     data = risk_trend_df)
summary(trend_geog_mdl2)

```

## Plot maps

``` {r maps}

mean_risk_df <- read_csv(file.path(dir_data, 'current_risk_by_loiczid.csv'), col_types = 'ddddd') %>%
  filter(n_spp >= 5)

mean_trend_df <- read_csv(file.path(dir_data, 'spp_trend_by_loiczid.csv'), col_types = 'ddd')

loiczid_rast <- raster(file.path(dir_spatial, 'loiczid_raster.tif'))

risk_rast <- subs(loiczid_rast, mean_risk_df, by = 'loiczid', which = 'log_mean_risk')

trend_rast <- subs(loiczid_rast, mean_trend_df, by = 'loiczid', which = 'mean_trend')

n_spp_rast <- subs(loiczid_rast, mean_risk_df, by = 'loiczid', which = 'n_spp')

```

### Mean spp risk (n_spp > 5)

``` {r}
plot(risk_rast)
```

### Mean spp trend (n_spp > 5)

``` {r}
plot(trend_rast)
```

### Number of species by risk

``` {r}
plot(n_spp_rast)
```


-----

``` {r prov_footer, results = 'asis'}
# prov_wrapup(commit_outputs = FALSE)
```

