---
title: 'Regress risk vs stressor groups'
author: "*Compiled on `r date()` by `r Sys.info()['user']`*"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    toc_float: yes
    number_sections: true
    theme: cerulean
    highlight: haddock
    includes: 
      in_header: '~/github/src/templates/ohara_hdr.html'
  pdf_document:
    toc: true
---

``` {r setup, echo = TRUE, message = FALSE, warning = FALSE}

knitr::opts_chunk$set(fig.width = 6, fig.height = 4, fig.path = 'Figs/',
                      echo = TRUE, message = FALSE, warning = FALSE)

library(raster)
library(sf)
library(fasterize)

source('https://raw.githubusercontent.com/oharac/src/master/R/common.R')

dir_git <- '~/github/spp_health_dists'

### goal specific folders and info
dir_data    <- file.path(dir_git, 'data')
dir_spatial <- file.path(dir_git, 'spatial')
dir_anx     <- file.path(dir_M, 'git-annex')
dir_o_anx   <- file.path(dir_O, 'git-annex/spp_health_dists')
# ### provenance tracking
# library(provRmd); prov_setup()

### support scripts
# source('https://raw.githubusercontent.com/oharac/src/master/R/rast_tools.R')) 
  ### raster plotting and analyzing scripts

```

# Summary

At the LOICZID raster scale (0.5° cells), compare ecosystem risk and trends to stressors layers.  Stressors are aggregated to LOICZID in a data_setup script.

# Data Sources

# Methods {.tabset}

For each stressors group, read in the various stressor layers from stressor_to_loiczid folder.  Join them and regress them against mean risk.  In each case, limit analysis to cells with at least 5 species indicated.

## Fisheries

``` {r log_fis_prs}

fis_stressor_files <- list.files('stressor_to_loiczid', 
                            pattern = 'demersal|pelagic|artisanal',
                            full.names = TRUE)

rm(fis_stressor_df)
for(stressor_file in fis_stressor_files) {  ### stressor_file <- fis_stressor_files[3]
  stressor_name <- basename(stressor_file) %>%
    str_replace('_simple.csv', '')
  
  # cat('Processing', stressor_name, '...\n')
  
  tmp <- read_csv(stressor_file, col_types = 'dddii') %>%
    select(-n_na) %>%
    setNames(c('loiczid', 
               paste0(stressor_name, '_mean'), paste0(stressor_name, '_var'), 
               paste0(stressor_name, '_zeros')))
  
  if(!exists('fis_stressor_df')) {
    fis_stressor_df <- tmp    ### create it the first time through the loop
  } else {
    fis_stressor_df <- fis_stressor_df %>%
      full_join(tmp, by = 'loiczid')     ### join it on subsequent times through the loop
  }
}

risk_df <- read_csv(file.path(dir_data, 'current_risk_by_loiczid.csv'),
                    col_types = 'ddddd')

risk_v_stressor_df <- full_join(risk_df, fis_stressor_df, by = 'loiczid') %>%
  filter(n_spp >= 5) ### leaves 145014 cells (out of 159535)
  # filter(n_spp >= 10) ### leaves 126466 cells

fis_stressor_mdl <- lm(log_mean_risk ~ artisanal_fishing_mean + 
                         demersal_destructive_fishing_mean + 
                         demersal_nondest_high_bycatch_mean +
                         pelagic_high_bycatch_mean +
                         pelagic_low_bycatch_mean,
                       data = risk_v_stressor_df)

summary(fis_stressor_mdl)

fis_stressor_mdl1 <- lm(mean_risk ~ artisanal_fishing_mean + 
                         demersal_destructive_fishing_mean + 
                         demersal_nondest_high_bycatch_mean +
                         pelagic_high_bycatch_mean +
                         pelagic_low_bycatch_mean,
                       data = risk_v_stressor_df)

summary(fis_stressor_mdl1)

```

## Pollution

``` {r pollution_prs}

poll_stressor_files <- list.files('stressor_to_loiczid', 
                            pattern = 'inorg|poll|plume',
                            full.names = TRUE)

rm(poll_stressor_df)
for(stressor_file in poll_stressor_files) {  ### stressor_file <- poll_stressor_files[3]
  stressor_name <- basename(stressor_file) %>%
    str_replace('_simple.csv', '')
  
  # cat('Processing', stressor_name, '...\n')
  
  tmp <- read_csv(stressor_file, col_types = 'dddii') %>%
    select(-n_na) %>%
    setNames(c('loiczid', 
               paste0(stressor_name, '_mean'), paste0(stressor_name, '_var'), 
               paste0(stressor_name, '_zeros')))
  
  if(!exists('poll_stressor_df')) {
    poll_stressor_df <- tmp    ### create it the first time through the loop
  } else {
    poll_stressor_df <- poll_stressor_df %>%
      full_join(tmp, by = 'loiczid')     ### join it on subsequent times through the loop
  }
}

risk_df <- read_csv(file.path(dir_data, 'current_risk_by_loiczid.csv'),
                    col_types = 'ddddd')

risk_v_stressor_df <- full_join(risk_df, poll_stressor_df, by = 'loiczid') %>%
  filter(n_spp >= 5)

poll_stressor_mdl <- lm(log_mean_risk ~ inorganic_mean + 
                          ocean_pollution_mean + 
                          plumes_fert_mean +
                          plumes_pest_mean,
                        data = risk_v_stressor_df)

summary(poll_stressor_mdl)

poll_stressor_mdl1 <- lm(mean_risk ~ inorganic_mean + 
                          ocean_pollution_mean + 
                          plumes_fert_mean +
                          plumes_pest_mean,
                        data = risk_v_stressor_df)

summary(poll_stressor_mdl1)

```

## Other non-climate

``` {r other_prs}

other_stressor_files <- list.files('stressor_to_loiczid', 
                            pattern = 'invasive|oil_rigs|night|pop|shipping',
                            full.names = TRUE)

rm(other_stressor_df)
for(stressor_file in other_stressor_files) {  ### stressor_file <- other_stressor_files[3]
  stressor_name <- basename(stressor_file) %>%
    str_replace('_simple.csv', '')
  
  # cat('Processing', stressor_name, '...\n')
  
  tmp <- read_csv(stressor_file, col_types = 'dddii') %>%
    select(-n_na) %>%
    setNames(c('loiczid', 
               paste0(stressor_name, '_mean'), paste0(stressor_name, '_var'), 
               paste0(stressor_name, '_zeros')))
  
  if(!exists('other_stressor_df')) {
    other_stressor_df <- tmp    ### create it the first time through the loop
  } else {
    other_stressor_df <- other_stressor_df %>%
      full_join(tmp, by = 'loiczid')     ### join it on subsequent times through the loop
  }
}

risk_df <- read_csv(file.path(dir_data, 'current_risk_by_loiczid.csv'),
                    col_types = 'ddddd')

risk_v_stressor_df <- full_join(risk_df, other_stressor_df, by = 'loiczid') %>%
  filter(n_spp >= 5)

other_stressor_mdl <- lm(log_mean_risk ~ invasives_mean + 
                           night_lights_mean + 
                           oil_rigs_mean +
                           population_mean + 
                           shipping_mean,
                         data = risk_v_stressor_df)

summary(other_stressor_mdl)

other_stressor_mdl1 <- lm(mean_risk ~ invasives_mean + 
                           night_lights_mean + 
                           oil_rigs_mean +
                           population_mean + 
                           shipping_mean,
                         data = risk_v_stressor_df)

summary(other_stressor_mdl1)

```

## Climate

``` {r cc_prs}

cc_stressor_files <- list.files('stressor_to_loiczid', 
                            pattern = 'slr|sst|uv|acid',
                            full.names = TRUE)

rm(cc_stressor_df)
for(stressor_file in cc_stressor_files) {  ### stressor_file <- cc_stressor_files[3]
  stressor_name <- basename(stressor_file) %>%
    str_replace('_simple.csv', '')
  
  # cat('Processing', stressor_name, '...\n')
  
  tmp <- read_csv(stressor_file, col_types = 'dddii') %>%
    select(-n_na) %>%
    setNames(c('loiczid', 
               paste0(stressor_name, '_mean'), paste0(stressor_name, '_var'), 
               paste0(stressor_name, '_zeros')))
  
  if(!exists('cc_stressor_df')) {
    cc_stressor_df <- tmp    ### create it the first time through the loop
  } else {
    cc_stressor_df <- cc_stressor_df %>%
      full_join(tmp, by = 'loiczid')     ### join it on subsequent times through the loop
  }
}

risk_df <- read_csv(file.path(dir_data, 'current_risk_by_loiczid.csv'),
                    col_types = 'ddddd')

lat_df <- read_csv(file.path(dir_data, 'latitude_lookup.csv'), col_types = 'dd')

risk_v_stressor_df <- full_join(risk_df, cc_stressor_df, by = 'loiczid') %>%
  left_join(lat_df, by = 'loiczid') %>%
  mutate(lat_abs = abs(lat)) %>%
  filter(n_spp >= 5)

cc_stressor_mdl <- lm(log_mean_risk ~ ocean_acidification_mean + ### on its own, R^2 = .0567
                    slr_mean + ### on its own, R^2 = .0198
                    sst_mean + ### on its own, R^2 = .0457
                    uv_mean,   ### on its own, R^2 = .2321, estimate = .149
                  data = risk_v_stressor_df)

summary(cc_stressor_mdl)

cc_stressor_mdl1 <- lm(mean_risk ~ ocean_acidification_mean + ### on its own, R^2 = .0567
                    slr_mean + ### on its own, R^2 = .0198
                    sst_mean + ### on its own, R^2 = .0457
                    uv_mean,   ### on its own, R^2 = .2321, estimate = .149
                  data = risk_v_stressor_df)

summary(cc_stressor_mdl1)


```

### What about UV and latitude?

Add latitude to the regression model for climate stressors especially UV; consider both latitude and absolute value of latitude (as a measure of symmetric distance toward poles)

``` {r cc_and_lat}

risk_v_stressor_df <- full_join(risk_df, cc_stressor_df, by = 'loiczid') %>%
  left_join(lat_df, by = 'loiczid') %>%
  mutate(lat_abs = abs(lat)) %>%
  filter(n_spp >= 5)

cc_stressor_mdl1 <- lm(log_mean_risk ~ ocean_acidification_mean + ### on its own, R^2 = .0567
                    slr_mean + ### on its own, R^2 = .0198
                    sst_mean + ### on its own, R^2 = .0457
                    uv_mean +  ### on its own, R^2 = .2321, estimate = .149
                    lat + lat_abs,
                  data = risk_v_stressor_df)
summary(cc_stressor_mdl1)

cc_stressor_mdl1 <- lm(mean_risk ~ ocean_acidification_mean + ### on its own, R^2 = .0567
                    slr_mean + ### on its own, R^2 = .0198
                    sst_mean + ### on its own, R^2 = .0457
                    uv_mean +  ### on its own, R^2 = .2321, estimate = .149
                    lat + lat_abs,
                  data = risk_v_stressor_df)
summary(cc_stressor_mdl1)

cc_stressor_mdl2 <- lm(log_mean_risk ~ uv_mean +   ### on its own, R^2 = .1572
                    # lat_abs +
                    # n_spp +
                    lat, ### on its own, R^2 = .0740
                  data = risk_v_stressor_df)

summary(cc_stressor_mdl2)

cc_stressor_mdl2 <- lm(mean_risk ~ uv_mean +   ### on its own, R^2 = .1572
                    # lat_abs +
                    # n_spp +
                    lat, ### on its own, R^2 = .0740
                  data = risk_v_stressor_df)

summary(cc_stressor_mdl2)

ggplot(risk_v_stressor_df, aes(x = uv_mean, y = log_mean_risk)) +
  geom_point(alpha = .01) +
  ggtheme_plot() +
  stat_smooth(method = 'lm')

ggplot(risk_v_stressor_df, aes(x = lat, y = uv_mean)) +
  geom_point(alpha = .01) +
  ggtheme_plot() +
  stat_smooth(method = 'lm')

ggplot(risk_v_stressor_df, aes(x = lat_abs, y = uv_mean)) +
  geom_point(alpha = .01) +
  ggtheme_plot() +
  stat_smooth(method = 'lm')

```

There seems to be a strong correlation between UV stressor (number of extreme UV events over five-year period) on its own and mean species extinction risk.  This seems to be separate from latitude-driven differences in UV events.  Other possibilities?

``` {r plot_uv_and_log_mean_risk}

log_mean_risk_df <- read_csv(file.path(dir_data, 'current_risk_by_loiczid.csv'), col_types = 'ddddd') %>%
  filter(n_spp >= 5)
uv_df <- read_csv(file.path(dir_git, 'stressor_to_loiczid/uv_simple.csv'), col_types = 'ddddd')

loiczid_rast <- raster(file.path(dir_spatial, 'loiczid_raster.tif'))

risk_rast <- subs(loiczid_rast, log_mean_risk_df, by = 'loiczid', which = 'log_mean_risk')

uv_rast <- subs(loiczid_rast, uv_df, by = 'loiczid', which = 'stressor_mean')

```

### Mean spp risk (n_spp > 5)

``` {r}
plot(risk_rast)
```

### UV stressor 

``` {r}
plot(uv_rast)
```

-----

``` {r prov_footer, results = 'asis'}
# prov_wrapup(commit_outputs = FALSE)
```

